---
title: "Data prep for Sóskuthy & Stuart-Smith 2018"
output:
  pdf_document: default
  html_notebook: default
---

## Vowel data

Let's import the data and relevant libraries first.

```{r}
library(tidyverse)
library(readr)
library(dplyr)
library(stringr)
library(lme4)
library(phonTools)


vowels <- c("boot", "cat", "cot", "face", "fleece", "goat", "strut")
files <- paste("data/midpoint_", toupper(vowels), ".csv", sep="")

for (f in 1:length(files)) {
  assign(vowels[f], read.csv(files[f]))
  assign(vowels[f], cbind(get(vowels[f]), data.frame(vowel=vowels[f])))
}
```

```{r}
# check colnames: print differences for each data frame

for (v1 in vowels) {
  for (v2 in vowels) {
    d <- setdiff(colnames(get(v1)), colnames(get(v2)))
    if (length(d) > 0) {
      cat(v1, "-", v2, ": ", paste(d, collapse=", "), "\n", sep="")
    }
  }
}

# identical, but no URL in fleece;
# face$Target.segments is all 1's (because of the celex transctiptions), 
# so should be converted to factor first;
# same for goat$Target.segments (all 5's)

face$Target.segments <- factor(face$Target.segments)
goat$Target.segments <- factor(goat$Target.segments)

# make list of data sets
all.list <- lapply(as.list(vowels), get, mode="list") # mode = "list" needed to make sure "cat" doesn't return function
# let's bind them together using bind_rows
all.vs <- bind_rows(all.list) # some columns coerced to character - should be OK

# fix "mother'd"
all.vs$Target.stress[all.vs$Target.stress=="'mVT"] <- "'mVD"
```

Now we'll get the following / preceding segment. These will be used as random intercepts in the model.

```{r}

# list of target segments (only used for extracting environment)
targets <- c(unique(all.vs$Target.segments), "U") # U (which doesn't contrast with u in Scottish varieties) missing from Target.segments but it's there in transcriptions
length(grep("[']", all.vs$Target.stress)) # all Target.stress values have '

# complicated regexp to get location of target segment in stressed syllable
# 2 actually means initial, since it's preceded by stress mark, so -1
segment.loc <- attr(regexpr(paste("'.*?([", paste(targets, collapse=""), "])", sep=""), all.vs$Target.stress, perl=T), "capture.start")[,1] - 1

# stringr to get location of target syllable in word
syllable.loc <- str_locate(all.vs$Target.CELEX.phonemes, fixed(str_sub(all.vs$Target.stress, 2, -1)))[,1] - 1

# so preceding is segment.loc + syllable.loc - 1, following is segment.loc + syllable.loc + 1
all.vs$preceding <- str_sub(all.vs$Target.CELEX.phonemes, segment.loc + syllable.loc - 1, segment.loc + syllable.loc - 1)
all.vs$following <- str_sub(all.vs$Target.CELEX.phonemes, segment.loc + syllable.loc + 1, segment.loc + syllable.loc + 1)
```

Let's create decade predictor.

```{r}
all.vs$decade <- str_sub(all.vs$Speaker, 1, 2)
all.vs$decade.fact <- factor(all.vs$decade, levels=c("70", "80", "90", "00")) # mind the ordering!
```

Vowel duration.

```{r}
all.vs$duration <- all.vs$Target.segments.end - all.vs$Target.segments.start
```

Let's exclude function words / words that are problematic for some reason.

```{r}
exclude.regs <- c(".*('ve|'d|'ll|~|'s|'t)","would","could","wouldnae","ye've","used","aboot","couldnae","oot","doon","should","that","have","has","what","because","not","says","st","don't","those","won't","just","wasnae","doesnae")
for (reg in exclude.regs) {
  all.vs <- filter(all.vs, !grepl(paste0("^", reg, "$"), Target.orthography))
}
```

Moving on to outliers & vowels with formant values that are fall outside reasonable bounds established on the basis of Hillenbrand et al. (1995).

```{r}
data(h95)

# ggplot(h95, aes(x=f1)) + facet_wrap(~type) + geom_density()
# ggplot(h95, aes(x=f2)) + facet_wrap(~type) + geom_density()
# ggplot(h95, aes(x=f3)) + facet_wrap(~type) + geom_density()
  
f_ranges <- h95 %>%
  group_by(type) %>%
  summarize(f1_lower=quantile(f1, 0.01),
            f1_upper=quantile(f1, 0.99),
            f2_lower=quantile(f2, 0.01),
            f2_upper=quantile(f2, 0.99),
            f3_lower=quantile(f3, 0.01),
            f3_upper=quantile(f3, 0.99)) %>%
  ungroup() %>%
  mutate(gender=as.character(recode(type, m="M", w="F"))) %>%
  dplyr::select(-type)

not_outlier <- function (x) {
  x > (quantile(x, 0.25) - 1.5*IQR(x)) & x < (quantile(x, 0.75) + 1.5*IQR(x))
}

all.vs_filtered <- all.vs %>%
  mutate(gender=as.character(gender)) %>%
  # filtering based on reasonable ranges / sex
  inner_join(f_ranges, by="gender") %>%
  filter(!is.na(F1.time_0.5) & !is.na(F2.time_0.5) & !is.na(F3.time_0.5)) %>%
  # marking clear f1/f2 outliers for each speaker as NAs
  group_by(Speaker, vowel) %>%
  mutate(f1=ifelse(F1.time_0.5 > f1_lower & F1.time_0.5 < f1_upper & not_outlier(F1.time_0.5), F1.time_0.5, NA),
         f2=ifelse(F2.time_0.5 > f2_lower & F2.time_0.5 < f2_upper & not_outlier(F2.time_0.5), F2.time_0.5, NA),
         f3=ifelse(F3.time_0.5 > f3_lower & F3.time_0.5 < f3_upper & not_outlier(F3.time_0.5), F3.time_0.5, NA)) %>%
  ungroup() %>%
  filter(!is.na(f1), !is.na(f2), !is.na(f3))
```

Fixing column labels, removing extraneous columns.

```{r}
all.vs_filtered <- all.vs_filtered %>%
  select(-SearchName, -Number, -Transcript, -Corpus, -Line, -LineEnd, -MatchId,
         -TargetId, -URL, -Text, -Target.stress.start, -Target.stress.end,
         -Target.transcript, -Target.transcript.end, -Target.transcript.start,
         -matches("_lower"), -matches("_upper"), -matches("[.]time_"), -Error)
```

And now save the final outcome.

```{r}
write_csv(all.vs_filtered, "data/vowels.csv")
```

We'll extract avg F2 and F3 values by speaker for use in regression models fitted to /r/. It would be useful to control for various things like vowel & context, so we'll fit a random effects model, with random intercepts by TargetOrthography, Vowel and Speaker. We'll compare the results with averages (mainly to show that the resulting values are *very* similar to the raw averages).

Calculate the averages first:

```{r}
# getting the averages
avg.f1 <- aggregate(F1.time_0.5 ~ Speaker, all.vs, mean)
avg.f2 <- aggregate(F2.time_0.5 ~ Speaker, all.vs, mean)
avg.f3 <- aggregate(F3.time_0.5 ~ Speaker, all.vs, mean)

# merging into a single data set + tidying
avg.fs <- merge(merge(avg.f1, avg.f2, by="Speaker"), avg.f3, by="Speaker")
avg.fs <- rename(avg.fs, speaker=Speaker, avg.f1=F1.time_0.5, avg.f2=F2.time_0.5, avg.f3=F3.time_0.5)
avg.fs$speaker <- as.character(avg.fs$speaker)
avg.fs$speaker[avg.fs$speaker=="80-O-f02-clbk"] <- "80-O-f02"
```

And now estimates from random models (i.e. controlling for speakers overusing specific vowels / words):

```{r}
rand.fs <- data.frame(speaker=as.character(unique(all.vs$Speaker)), stringsAsFactors = F)
for (f in 1:3) {
  rand.formula <- as.formula(paste0("F", f, ".time_0.5 ~ (1 | Speaker) + (1 | Target.orthography) + (1 | vowel)"))
  rand.model <- lmer(rand.formula, data=all.vs)
  per.speaker.intercepts <- ranef(rand.model)$Speaker
  colnames(per.speaker.intercepts) <- paste0("rand.f", f)
  per.speaker.intercepts$speaker <- rownames(per.speaker.intercepts)
  rand.fs <- merge(rand.fs, per.speaker.intercepts, by="speaker", stringsAsFactors=F)
}
rand.fs$speaker[rand.fs$speaker=="80-O-f02-clbk"] <- "80-O-f02"
```

And now merging the two sets of estimates and comparing them.

```{r}
aggr.fs <- merge(avg.fs, rand.fs, by="speaker")

plot(aggr.fs$avg.f1, aggr.fs$rand.f1)
plot(aggr.fs$avg.f2, aggr.fs$rand.f2)
plot(aggr.fs$avg.f3, aggr.fs$rand.f3) 
```

## /r/

```{r}
R <- readRDS("data/R_fixed.rds")
R <- merge(R, aggr.fs, by="speaker")
R$freq.log <- log(R$freq+1)

write_csv(R, "data/R.csv")
```