---
title: Plots for `Voice quality and coda /r/ in Glasgow English in the early 20th
  century'
output:
  html_document:
    df_print: paged
  html_notebook: default
  pdf_document: default
---

This R-markdown file can be used to recreate the plots in our paper titled "Voice quality and coda /r/ in Glasgow English in the early 20th century". The plots are presented in the same order as they are in the paper, and the section numbering also follows our paper (which means that it actually starts at 4.1, which is the first results section in the paper).

We start by importing all relevant libraries and the data. The details of the data processing (e.g. exclusion of outliers) are shown in a separate file titled *data_prep.Rmd*.

Let's import the data and relevant libraries first. Note that we have saved our statistical models as RDS files, which allows the user to save time by not having to refit them.

```{r}
library(lme4)
library(mgcv)
library(itsadug)
library(rms)
library(tidyverse)
library(stringr)
library(effects)
library(here)
library(afex)

# vowel data
vowels <- read_csv("data/vowels.csv")
# formatting for mixed effects models
vowels$Speaker <- as.factor(vowels$Speaker)
vowels$Target.orthography <- as.factor(vowels$Target.orthography)
vowels$vowel <- as.factor(vowels$vowel)
vowels$decade <- ifelse(vowels$decade=="00", 2000, as.numeric(vowels$decade) + 1900)
vowels$decade.fact <- factor(vowels$decade.fact, levels=c("70","80","90","00"))


# /r/ data

R <- read_csv("data/R.csv") # F/M confuse read_csv...
coltyps <- spec(R)
coltyps$cols$gender <- col_character()
# and again!
R <- read_csv("data/R.csv", col_types=coltyps)

# some changes to variables to make them work with GAMMs
R <- R %>%
  mutate(stress = as.ordered(stress),
         start.event = (measurement.no==0),
         foll.broad.f3 = as.factor(foll.broad.f3),
         speaker = as.factor(speaker),
         word = as.factor(word))
contrasts(R$stress) <- "contr.treatment"

# separate data sets for males and females
R.m <- R %>%
  filter(gender=="M")
R.f <- R %>%
  filter(gender=="F")

# setting up variables for analysis via GAMMs
R$trans.broad <- recode(R$trans, wa="a")
R$trans.broad.fact <- as.ordered(R$trans.broad)
contrasts(R$trans.broad.fact) <- "contr.treatment"
R$gender.ordered <- as.ordered(R$gender)
contrasts(R$gender.ordered) <- "contr.treatment"


# auditory R data -- create data set with unique realisations
R.aud <- R[!(R$trans %in% c("", "?r")) & R$measurement.no==0,]
# only a few weakened approximants, so fold these into other categories
R.aud <- R %>%
  filter(!(trans %in% c("", "?r")), 
         !is.na(trans),
         measurement.no==0) %>%
  mutate(trans.broad=recode(trans, wa="a"),
         gender=recode(gender, M="male", `F`="female"))

# auditory voice quality data
# getting f3 avgs
avg.f3 <- unique(select(R, speaker, avg.f3))
# reading & formatting data
R.vq <- read_csv("data/auditoryvq.csv") %>%
  rename(speaker="speaker code",
         advanced.tip.blade="advanced tip/blade",
         tongue.body.height="tongue body height",
         tongue.body.front.backness="tongue body front-backness") %>%
  # join with avg.f3!
  inner_join(avg.f3, by="speaker") %>%
  mutate(gender=as.factor(ifelse(!is.na(str_match(speaker, "-m0")), "male", "female")),
         decade=unlist(substr(speaker, 1, 2)),
         decade=as.numeric(recode(decade, `00`="100")) + 1900)

```

# 4.1: Changes to coda /r/ -- Take 1: Acoustic results

## Males without baseline control

Creating figure 2 bottom panel showing the model predictions for males.

```{r}
# importing statistical model
R.m.mod.AR <- readRDS("models/R.m.mod.AR")

# extracting model predictions (using plot_smooth from itsadug)
R.m.full.plot <- plot_smooth(R.m.mod.AR, view="measurement.no", plot_all="decade", 
            cond=list(stress="full"), rm.ranef=T, n.grid=100)$fv
R.m.schwa.plot <- plot_smooth(R.m.mod.AR, view="measurement.no", plot_all="decade", 
            cond=list(stress="schwa"), rm.ranef=T, n.grid=100)$fv
R.m.plot <- rbind(R.m.full.plot, R.m.schwa.plot)

# formatting decade predictor
R.m.plot$decade <- factor(R.m.plot$decade, levels=c("1970","1980","1990","2000"))

# creating the plot
ggplot(R.m.plot, aes(x=measurement.no, y=fit, group=decade, col=decade, lty=decade)) +
  geom_line(lwd=1) +
  facet_grid(.~stress) +
  geom_ribbon(aes(ymin=ll, ymax=ul, group=decade), col=NA,col="grey", alpha=0.1) +
  scale_color_manual(values=c("orange","darkorange3","deepskyblue1","deepskyblue4"),
                     name="decade of\nrecording") +
  scale_linetype_discrete(name="decade of\nrecording") +
  scale_x_continuous(name="measurement point", limits = c(0, 10), breaks=seq(0,10,2)) +
  scale_y_continuous(name="F3 (Hz)", limits=c(2000, 3050)) +
  theme_bw() +
  theme(axis.title = element_text(size=14), 
        axis.text = element_text(size=12),
        legend.title = element_text(size=14, face="bold"),
        legend.text=element_text(size=12),
        plot.title=element_text(size=14, face="bold"),
        panel.grid=element_blank(),
        strip.text=element_text(size=12)) +
  ggtitle("F3 changes for /r/ in males (not controlling for baseline F3)")
#ggsave("graphs/r_male_f3_change_no_control.pdf", width=8, height=4)
```



## Females without baseline control

Creating figure 2 top panel showing the model predictions for females.

```{r}
# importing statistical model
R.f.mod.AR <- readRDS("models/R.f.mod.AR")

# extracting model predictions (using plot_smooth from itsadug)
R.f.full.plot <- plot_smooth(R.f.mod.AR, view="measurement.no", plot_all="decade", 
            cond=list(stress="full"), rm.ranef=T, n.grid=100)$fv
R.f.schwa.plot <- plot_smooth(R.f.mod.AR, view="measurement.no", plot_all="decade", 
            cond=list(stress="schwa"), rm.ranef=T, n.grid=100)$fv
R.f.plot <- rbind(R.f.full.plot, R.f.schwa.plot)

# formatting decade predictor
R.f.plot$decade <- factor(R.f.plot$decade, levels=c("1970","1980","1990","2000"))
  
# creating plot
ggplot(R.f.plot, aes(x=measurement.no, y=fit, group=decade, col=decade, lty=decade)) +
  geom_line(lwd=1) +
  facet_grid(.~stress) +
  geom_ribbon(aes(ymin=ll, ymax=ul, group=decade), col=NA,col="grey", alpha=0.1) +
  scale_color_manual(name="decade of\nrecording", values=c("orange","darkorange3","deepskyblue1","deepskyblue4")) +
  scale_linetype_discrete(name="decade of\nrecording") +
  scale_x_continuous(name="measurement point", limits = c(0, 10), breaks=seq(0,10,2)) +
  scale_y_continuous(name="F3 (Hz)", limits=c(2000, 3050)) +
  theme_bw() +
  theme(axis.title = element_text(size=14), 
        axis.text = element_text(size=12),
        legend.title = element_text(size=14, face="bold"),
        legend.text=element_text(size=12),
        plot.title=element_text(size=14, face="bold"),
        panel.grid=element_blank(),
        strip.text=element_text(size=12)) +
  ggtitle("F3 changes for /r/ in females (not controlling for baseline F3)")
#ggsave("graphs/r_female_f3_change_no_control.pdf", width=8, height=4)
```

# 4.2: Changes to coda /r/ -- Take 1: Auditory results

Creating figure 3.

```{r}
# importing model
R.shapes.mod.AR <- readRDS("models/R.shapes.mod.AR")

# extracting model predictions
R.shapes <- plot_smooth(R.shapes.mod.AR, view="measurement.no", plot_all="trans.broad.fact",
                        cond=list(gender="F"), rm.ranef=T, rug=F)[[1]]

# releveling, meaningful names for realisations
R.shapes$trans.broad.fact <- factor(R.shapes$trans.broad.fact, levels=c("o","i","wt","a","t","r"))
R.shapes <- R.shapes %>% 
  mutate(realisation=recode(trans.broad.fact,
                            o="zero",
                            i="intermediate",
                            wt="weakened tap",
                            a="approximant",
                            t="full tap",
                            r="trill"),
         realisation=factor(realisation, 
                            levels=c("zero",
                                     "intermediate",
                                     "weakened tap",
                                     "approximant",
                                     "full tap",
                                     "trill"))
  )
# excluding trills
R.shapes <- filter(R.shapes, trans.broad.fact != "r")

# creating graph
ggplot(R.shapes, aes(x=measurement.no, y=fit, col=realisation, lty=realisation)) +
  geom_ribbon(aes(ymin=ll, ymax=ul, group=trans.broad.fact), alpha=0.05, colour=NA) +
  geom_line(lwd=1) +
  scale_linetype_manual(values = c(3,4,2,
                              5,1),
                     name="/r/ realisation") +
  scale_colour_manual(values = c("orange","darkorange1",
                               "darkorange3","deepskyblue1",
                               "deepskyblue3"),
                      name="/r/ realisation") +
  scale_x_continuous(name="measurement point", limits = c(0, 10), breaks=seq(0,10,2)) +
  scale_y_continuous(name="F3 (Hz)", limits=c(2450,2950)) +
  theme_bw() +
  theme(axis.title = element_text(size=14), 
        axis.text = element_text(size=12),
        legend.title = element_text(size=14, face="bold"),
        legend.text=element_text(size=12),
        plot.title=element_text(size=14, face="bold"),
        panel.grid=element_blank(),
        strip.text=element_text(size=12)) +
  ggtitle("Acoustic correlates of different /r/ realisations in F3")
#ggsave("graphs/r_acoustic_correlates.pdf", width=6, height=4)
```

We now create figure 4, that is, the proportions of different realisations as a function of time.

```{r}
# calculating proportions
R.props.dec <- R.aud %>% 
  count(decade, gender, trans.broad) %>% 
  group_by(decade, gender) %>%
  mutate(prop = n/sum(n)) %>%
  ungroup()

# releveling
R.props.dec$trans.broad <- factor(R.props.dec$trans.broad, levels=c("o","i","wt","a","t","r"))

# creating the graph
ggplot(R.props.dec, aes(x=decade, fill=trans.broad)) + 
  geom_bar(aes(y=prop), stat="identity", position="stack") + 
  facet_wrap(~gender) + 
  scale_fill_manual(values = c("orange","darkorange1",
                               "darkorange3","deepskyblue1",
                               "deepskyblue3","deepskyblue4"),
                     name="/r/ realisation", 
                    labels=c("zero","intermediate",
                             "weakened tap","approximant",
                             "tap","trill")) +
  scale_y_continuous(labels = scales::percent, name="percentage of realisations") +
  scale_x_continuous(name = "decade of recording") +
  theme_bw() +
  theme(axis.title = element_text(size=14), 
        axis.text = element_text(size=12),
        legend.title = element_text(size=14, face="bold"),
        legend.text=element_text(size=12),
        plot.title=element_text(size=14, face="bold"),
        panel.grid=element_blank(),
        strip.text=element_text(size=12)) +
ggtitle("Changes in proportions of auditory /r/ variants")
#ggsave("graphs/r_auditory.pdf", width=8, height=4)
```

# 5.1 Acoustic results for voice quality

Prediction plot for the paper, i.e. figure 5:

```{r}
f3.mod.full <- readRDS("models/f3.mod.full")

# extracting predictions from the model
new.dat <- expand.grid(gender=unique(vowels$gender),
                       decade.fact=levels(vowels$decade.fact),
                       duration=median(vowels$duration),
                       vowel=unique(vowels$vowel)[1])
new.dat$f3 <- 0
new.dat$f3 <- predict(f3.mod.full, new.dat, re.form=NA)
mm <- model.matrix(terms(f3.mod.full),new.dat) # for confidence intervals
pvar1 <- diag(mm %*% tcrossprod(vcov(f3.mod.full),mm))

new.dat <- data.frame(
    new.dat,
    lower = new.dat$f3-2*sqrt(pvar1),
    upper = new.dat$f3+2*sqrt(pvar1)
)

# creating the graph

mean_fun <- function(x){
  return(data.frame(y = mean(x), label = paste0("  ", round(mean(x)), " Hz")))
}

ggplot(new.dat, aes(x=decade.fact, y=f3, col=gender)) +
  geom_violin(data=vowels, aes(fill=gender, col=NA), alpha=0.2, show.legend=F) +
  geom_point(size=2, position=position_dodge(width=0.9)) +
  geom_errorbar(aes(ymin=lower, ymax=upper), width=0.25, position=position_dodge(width=0.9)) +
  stat_summary(fun.data = mean_fun, geom = "text", size=4, position=position_dodge(0.9), hjust=0, show.legend=F) +
  scale_color_manual(name="gender",breaks=c("F","M"), labels=c("female", "male"), 
                     values=c("deepskyblue4","orange")) +
  scale_fill_manual(name="gender",breaks=c("F","M"), labels=c("female", "male"), 
                     values=c("deepskyblue4","orange"), guide=F) +
  scale_x_discrete(name="decade of recording", labels=c(1970,1980,1990,2000), expand=expand_scale(add=c(0.05,1))) +
  scale_y_continuous(name="F3") +
  theme_bw() +
  theme(axis.title = element_text(size=14), 
        axis.text = element_text(size=12),
        legend.title = element_text(size=14, face="bold"),
        legend.text=element_text(size=12),
        plot.title=element_text(size=14, face="bold"),
        panel.grid=element_blank()) +
  ggtitle("F3 changes for vowels (with model predictions)")
#ggsave("graphs/vowels_f3_change.pdf", width=8, height=4)
  
```


# 5.2 Auditory results for voice quality

Create figure 6.

```{r}
# getting counts of different tongue body height values per decade
R.vq.count <- dplyr::count(R.vq, tongue.body.height, decade)
# arranging by decade / tongue body height
R.vq.ordered <- R.vq %>% 
  arrange(decade, tongue.body.height)

# creating graph
ggplot(R.vq.ordered, aes(x=factor(decade), y=tongue.body.height)) +
  geom_point(x=1,y=1, aes(colour="Female")) + # this is a hack for getting the legend right 
  geom_point(x=1,y=1, aes(colour="Male")) + # this is a hack for getting the legend right 
  scale_color_manual(name="Gender", values=c("deepskyblue4","orange"), breaks=c("Female","Male")) +
  geom_dotplot(binaxis="y", stackdir="center", binpositions="all",
               fill=ifelse(R.vq.ordered$gender=="female", "deepskyblue4","orange"),
               col=NA, dotsize=1.5) +
  stat_summary(fun.y=mean, fun.ymin=mean, fun.ymax=mean, geom="crossbar", aes(group=decade), width=0.4, col="grey") +
  scale_x_discrete(name="Decade of recording") +
  scale_y_continuous(name="Tongue body height") +
  guides(colour = guide_legend(override.aes = list(size = 4))) +
  theme_bw() +
  theme(axis.title = element_text(size=14), 
        axis.text = element_text(size=12),
        legend.title = element_text(size=14, face="bold"),
        legend.text=element_text(size=12),
        plot.title=element_text(size=14, face="bold"),
        panel.grid=element_blank(),
        strip.text=element_text(size=12)) +
  ggtitle("Changes in tongue body height over time")
# ggsave("graphs/tongue_body_height.pdf", width=6, height=4)
```

# Changes to Coda R: take 2 -- acoustic analysis

## Males, take 2 (with baseline control)

Creating the bottom panel of figure 7.

```{r}
# importing model
R.m.baseline.mod.AR <- readRDS("models/R.m.baseline.mod.AR")

# extracting predictions, formatting decade
R.m.baseline.full.plot <- plot_smooth(R.m.baseline.mod.AR, view="measurement.no", plot_all="decade", 
            cond=list(stress="full"), rm.ranef=T, n.grid=100)$fv
R.m.baseline.schwa.plot <- plot_smooth(R.m.baseline.mod.AR, view="measurement.no", plot_all="decade", 
            cond=list(stress="schwa"), rm.ranef=T, n.grid=100)$fv
R.m.baseline.plot <- rbind(R.m.baseline.full.plot, R.m.baseline.schwa.plot)
R.m.baseline.plot$decade <- factor(R.m.baseline.plot$decade, levels=c("1970","1980","1990","2000"))

# create graph
ggplot(R.m.baseline.plot, aes(x=measurement.no, y=fit, group=decade, col=decade, lty=decade)) +
  geom_line(lwd=1) +
  facet_grid(.~stress) +
  geom_ribbon(aes(ymin=ll, ymax=ul, group=decade), col=NA,col="grey", alpha=0.1) +
  scale_color_manual(name="decade of\nrecording", values=c("orange","darkorange3","deepskyblue1","deepskyblue4")) +
  scale_linetype_discrete(name="decade of\nrecording") +
  scale_x_continuous(name="measurement point", limits = c(0, 10), breaks=seq(0,10,2)) +
  scale_y_continuous(name="F3 (Hz)", limits=c(2000, 3050)) +
  theme_bw() +
  theme(axis.title = element_text(size=14), 
        axis.text = element_text(size=12),
        legend.title = element_text(size=14, face="bold"),
        legend.text=element_text(size=12),
        plot.title=element_text(size=14, face="bold"),
        panel.grid=element_blank(),
        strip.text=element_text(size=12)) +
  ggtitle("F3 changes for /r/ in males (controlling for baseline F3)")
#ggsave("graphs/r_male_f3_change_control.pdf", width=8, height=4)
```

## Females, take 2 (with baseline control)

Creating the top panel of figure 7.

```{r}
# importing model
R.f.baseline.mod.AR <- readRDS("models/R.f.baseline.mod.AR")

# extracting predictions, formatting decades
R.f.baseline.full.plot <- plot_smooth(R.f.baseline.mod.AR, view="measurement.no", plot_all="decade", 
            cond=list(stress="full"), rm.ranef=T, n.grid=100)$fv
R.f.baseline.schwa.plot <- plot_smooth(R.f.baseline.mod.AR, view="measurement.no", plot_all="decade", 
            cond=list(stress="schwa"), rm.ranef=T, n.grid=100)$fv
R.f.baseline.plot <- rbind(R.f.baseline.full.plot, R.f.baseline.schwa.plot)
R.f.baseline.plot$decade <- factor(R.f.baseline.plot$decade, levels=c("1970","1980","1990","2000"))

# creating graph
ggplot(R.f.baseline.plot, aes(x=measurement.no, y=fit, group=decade, col=decade, lty=decade)) +
  geom_line(lwd=1) +
  facet_grid(.~stress) +
  geom_ribbon(aes(ymin=ll, ymax=ul, group=decade), col=NA,col="grey", alpha=0.1) +
  scale_color_manual(name="decade of\nrecording", values=c("orange","darkorange3","deepskyblue1","deepskyblue4")) +
  scale_linetype_discrete(name="decade of\nrecording") +
  scale_x_continuous(name="measurement point", limits = c(0, 10), breaks=seq(0,10,2)) +
  scale_y_continuous(name="F3 (Hz)", limits=c(2000, 3050)) +
  theme_bw() +
  theme(axis.title = element_text(size=14), 
        axis.text = element_text(size=12),
        legend.title = element_text(size=14, face="bold"),
        legend.text=element_text(size=12),
        plot.title=element_text(size=14, face="bold"),
        panel.grid=element_blank(),
        strip.text=element_text(size=12)) +
  ggtitle("F3 changes for /r/ in females (controlling for baseline F3)")
#ggsave("graphs/r_female_f3_change_control.pdf", width=8, height=4)
```





