---
title: "F3 changes in Glasgow vowels / r"
output:
  pdf_document: default
  html_notebook: default
---

## Vowel data

# Data wrangling

Let's import the data and relevant libraries first.

```{r}
library(lme4)
library(mgcv)
library(itsadug)
library(rms)
library(tidyverse)
library(stringr)
library(effects)
library(here)
library(afex)

refit_please <- F # please don't refit all the models - just load them from RDS files
vowels <- read_csv("data/vowels.csv")
```


# Data visualisation

Here is a summary of the data.

```{r}
summary(vowels)
```

Let's go through the list:

- Speaker as random intercept
- gender as fixed pred, possibly in interaction with decade
- for simplicity, decade as linear?
- vowel duration with random slopes by vowel
- random intercepts by vowel, following, preceding

Would be nice to create some raw plots. How about boxplots of F2 / F3 / vowel / gender?

```{r}
vowels$decade.fact <- factor(vowels$decade.fact, levels=c("70","80","90","00"))
ggplot(vowels, aes(x=decade.fact, y=f1)) +
  facet_grid(gender ~ vowel) +
  geom_boxplot()
ggplot(vowels, aes(x=decade.fact, y=f2)) +
  facet_grid(gender ~ vowel) +
  geom_boxplot()
ggplot(vowels, aes(x=decade.fact, y=f3)) +
  facet_grid(gender ~ vowel) +
  geom_boxplot()
```

This is *awesome*! F1/F2 shows all kinds of changes / stable distributions, perhaps unsurprisingly. However, F3 rises very consistently across all vowels for both genders.

# Modelling F1/F2/F3

We'll test this using linear mixed models. The dependent variables will be F1/F2/F3. We'll run the models one by one.

Data formatting:

```{r, cache=T}
vowels$Speaker <- as.factor(vowels$Speaker)
vowels$Target.orthography <- as.factor(vowels$Target.orthography)
vowels$vowel <- as.factor(vowels$vowel)
vowels$decade <- ifelse(vowels$decade=="00", 2000, as.numeric(vowels$decade) + 1900)
vowels$decade.fact <- factor(vowels$decade.fact, levels=c("70","80","90","00"))
```

First F1:

```{r, cache=T}
# this is the model that we would like to have - but it
# does not converge due to a singularity involving 
# the decade by vowel random slope

# f1.mod.full <- lmer(f1 ~ gender * decade.fact +
#                           scale(duration) +
#                             (1 + decade.fact | vowel) +
#                             (1 | Target.orthography) +
#                             (1 | Speaker),
#               data=vowels, REML=F)

# these are two alternative models; model 1 forces the decade
# effect to be linear

if (refit_please) {
  f1.mod.full.1 <- lmer(f1 ~ gender * scale(decade) + 
                           scale(duration) +
                               (1 + scale(decade) | vowel) +
                               (1 | Target.orthography) +
                               (1 | Speaker),
                 data=vowels, REML=F)
  # saveRDS(f1.mod.full.1, "models/f1.mod.full.1")
  
  # model 2 has no by-vowel random slope for decade (but the decade
  # effect is allowed to be non-linear)
  
  f1.mod.full.2 <- lmer(f1 ~ gender * decade.fact + 
                           scale(duration) +
                               (1 | vowel) +
                               (1 | Target.orthography) +
                               (1 | Speaker),
                 data=vowels, REML=F)
  # saveRDS(f1.mod.full.2, "models/f1.mod.full.2")
  #summary(f3.mod.full)
  
  # neither model shows a significant effect of decade:
  # (and note that model 2 is expected to be anti-conservative with
  #  respect to decade!)
  
  f1.mod.comp.1 <- lmer(f1 ~ gender + scale(duration) +
                               (1 + scale(decade) | vowel) +
                               (1 | Target.orthography) +
                               (1 | Speaker),
                 data=vowels, REML=F)
  # saveRDS(f1.mod.comp.1, "models/f1.mod.comp.1")
  f1.mod.comp.2 <- lmer(f1 ~ gender + scale(duration) +
                               (1 | vowel) +
                               (1 | Target.orthography) +
                               (1 | Speaker),
                 data=vowels, REML=F)
  # saveRDS(f1.mod.comp.2, "models/f1.mod.comp.2")
}

f1.mod.full.1 <- readRDS("models/f1.mod.full.1")
f1.mod.full.2 <- readRDS("models/f1.mod.full.2")
f1.mod.comp.1 <- readRDS("models/f1.mod.comp.1")
f1.mod.comp.2 <- readRDS("models/f1.mod.comp.2")
anova(f1.mod.full.1, f1.mod.comp.1)
anova(f1.mod.full.2, f1.mod.comp.2)
```

```{r, cache=T}
# this is the model that we would like to have - but it
# does not converge due to a singularity involving 
# the decade by vowel random slope

# f2.mod.full <- lmer(f2 ~ gender * decade.fact +
#                           scale(duration) +
#                             (1 + decade.fact | vowel) +
#                             (1 | Target.orthography) +
#                             (1 | Speaker),
#               data=vowels, REML=F)

# these are two alternative models; model 1 forces the decade
# effect to be linear

if (refit_please) {
  f2.mod.full.1 <- lmer(f2 ~ gender * scale(decade) + 
                           scale(duration) +
                               (1 + scale(decade) | vowel) +
                               (1 | Target.orthography) +
                               (1 | Speaker),
                 data=vowels, REML=F)
  # saveRDS(f2.mod.full.1, "models/f2.mod.full.1")
  
  # model 2 has no by-vowel random slope for decade (but the decade
  # effect is allowed to be non-linear)
  
  f2.mod.full.2 <- lmer(f2 ~ gender * decade.fact + 
                           scale(duration) +
                               (1 | vowel) +
                               (1 | Target.orthography) +
                               (1 | Speaker),
                 data=vowels, REML=F)
  # saveRDS(f2.mod.full.2, "models/f2.mod.full.2")
  #summary(f3.mod.full)
  
  # neither model shows a significant effect of decade:
  # (and note that model 2 is expected to be anti-conservative with
  #  respect to decade!)
  
  f2.mod.comp.1 <- lmer(f2 ~ gender + scale(duration) +
                               (1 + scale(decade) | vowel) +
                               (1 | Target.orthography) +
                               (1 | Speaker),
                 data=vowels, REML=F)
  # saveRDS(f2.mod.comp.1, "models/f2.mod.comp.1")
  f2.mod.comp.2 <- lmer(f2 ~ gender + scale(duration) +
                               (1 | vowel) +
                               (1 | Target.orthography) +
                               (1 | Speaker),
                 data=vowels, REML=F)
  # saveRDS(f2.mod.comp.2, "models/f2.mod.comp.2")
}

f2.mod.full.1 <- readRDS("models/f2.mod.full.1")
f2.mod.comp.1 <- readRDS("models/f2.mod.comp.1")
f2.mod.full.2 <- readRDS("models/f2.mod.full.2")
f2.mod.comp.2 <- readRDS("models/f2.mod.comp.2")

anova(f2.mod.full.1, f2.mod.comp.1)
anova(f2.mod.full.2, f2.mod.comp.2)
```

F2 does not show a significant change.

Now for F3:

```{r}
if (refit_please) {
  f3.mod.full <- lmer(f3 ~ gender * decade.fact + 
                           scale(duration) +
                               (1 + decade.fact | vowel) +
                               (1 | Target.orthography) +
                               (1 | Speaker),
                 data=vowels, REML=F)
  # saveRDS(f3.mod.full, "models/f3.mod.full")
  #summary(f3.mod.full)
  
  
  f3.mod.comp <- lmer(f3 ~ gender + scale(duration) +
                               (1 + decade.fact | vowel) +
                               (1 | Target.orthography) +
                               (1 | Speaker),
                 data=vowels, REML=F)
  # saveRDS(f3.mod.comp, "models/f3.mod.comp")
}

f3.mod.full <- readRDS("models/f3.mod.full")
f3.mod.comp <- readRDS("models/f3.mod.comp")

anova(f3.mod.full, f3.mod.comp)
```

This is so cool! F3 clearly rises very significantly over time. Here's a plot:

```{r}
new.dat <- expand.grid(gender=unique(vowels$gender),
                       decade.fact=levels(vowels$decade.fact),
                       duration=median(vowels$duration),
                       #following=levels(vowels$following)[1],
                       #preceding=levels(vowels$preceding)[1],
                       vowel=unique(vowels$vowel)[1])
new.dat$f3 <- 0
mm <- model.matrix(terms(f3.mod.full),new.dat)
new.dat$f3 <- predict(f3.mod.full, new.dat, re.form=NA)
pvar1 <- diag(mm %*% tcrossprod(vcov(f3.mod.full),mm))

new.dat <- data.frame(
    new.dat,
    lower = new.dat$f3-2*sqrt(pvar1),
    upper = new.dat$f3+2*sqrt(pvar1)
)

ggplot(new.dat, aes(x=decade.fact, y=f3, col=gender)) +
  #geom_ribbon(aes(ymin=lower, ymax=upper, group=gender), col="grey", alpha=0.3) +
  geom_point(size=2) +
  geom_errorbar(aes(ymin=lower, ymax=upper), width=0.25)

## graph for abstract:

#med_fun <- function(x){
#  return(data.frame(y = median(x), label = paste0(median(x), "%")))
#}

mean_fun <- function(x){
  return(data.frame(y = mean(x), label = paste0("  ", round(mean(x)), " Hz")))
}

ggplot(new.dat, aes(x=decade.fact, y=f3, col=gender)) +
  #geom_ribbon(aes(ymin=lower, ymax=upper, group=gender), col="grey", alpha=0.3) +
  geom_violin(data=vowels, aes(fill=gender, col=NA), alpha=0.2, show.legend=F) +
  geom_point(size=2, position=position_dodge(width=0.9)) +
  geom_errorbar(aes(ymin=lower, ymax=upper), width=0.25, position=position_dodge(width=0.9)) +
  stat_summary(fun.data = mean_fun, geom = "text", size=4, position=position_dodge(0.9), hjust=0, show.legend=F) +
  scale_color_manual(name="gender",breaks=c("F","M"), labels=c("female", "male"), 
                     values=c("deepskyblue4","orange")) +
  scale_fill_manual(name="gender",breaks=c("F","M"), labels=c("female", "male"), 
                     values=c("deepskyblue4","orange"), guide=F) +
  scale_x_discrete(name="decade of recording", labels=c(1970,1980,1990,2000), expand=expand_scale(add=c(0.05,1))) +
  scale_y_continuous(name="F3") +
  theme_bw() +
  theme(axis.title = element_text(size=14), 
        axis.text = element_text(size=12),
        legend.title = element_text(size=14, face="bold"),
        legend.text=element_text(size=12),
        plot.title=element_text(size=14, face="bold"),
        panel.grid=element_blank()) +
  ggtitle("F3 changes for vowels (with model predictions)")
ggsave("graphs/vowels_f3_change.pdf", width=8, height=4)
  
```


## Acoustic analysis of /r/

Data import / wrangling:

```{r}
library(mgcv)
library(itsadug)
# F/M confuse read_csv...
R <- read_csv("data/R.csv")
coltyps <- spec(R)
coltyps$cols$gender <- col_character()
# and again!
R <- read_csv("data/R.csv", col_types=coltyps)

R <- R %>%
  mutate(stress = as.ordered(stress),
         start.event = (measurement.no==0),
         foll.broad.f3 = as.factor(foll.broad.f3),
         speaker = as.factor(speaker),
         word = as.factor(word))
contrasts(R$stress) <- "contr.treatment"

R.m <- R %>%
  filter(gender=="M")
R.f <- R %>%
  filter(gender=="F")
```

# Males without baseline control

Analysing males first, no baseline F3. Fitting:

- model without AR for estimating rho
- full model with AR
- nested model with AR (rho from full model)

```{r}
if (refit_please) {
  system.time(
    R.m.mod.no.AR <- bam(f3 ~ stress +
                       # level 1 smooth terms
                          s(measurement.no) +
                          s(decade, k=4) +
                          s(duration) +
                          s(preceding.front, k=3) +
                       # level 2: by terms
                          s(measurement.no, by=stress) +
                          s(decade, k=4, by=stress) +
                       # level 2: 2d smooths
                          ti(measurement.no, decade, k=c(10,4)) +
                          ti(measurement.no, duration) +
                          ti(measurement.no, preceding.front, k=c(10,3)) +
                       # level 3: 2d smooth by terms
                          ti(measurement.no, decade, k=c(10,4), by=stress) +
                       # random smooths
                          s(measurement.no, foll.broad.f3, bs="fs", m=1, k=4) +
                          s(measurement.no, speaker, bs="fs", m=1, k=4) +
                          s(measurement.no, word, bs="fs", m=1, k=4),
                      dat=R.m)
  )
  # saveRDS(R.m.mod.no.AR, "models/R.m.mod.no.AR")
  
  R.m.rho <- start_value_rho(R.m.mod.no.AR)
  
  system.time(
    R.m.mod.AR <- bam(f3 ~ stress +
                       # level 1 smooth terms
                          s(measurement.no) +
                          s(decade, k=4) +
                          s(duration) +
                          s(preceding.front, k=3) +
                       # level 2: by terms
                          s(measurement.no, by=stress) +
                          s(decade, k=4, by=stress) +
                       # level 2: 2d smooths
                          ti(measurement.no, decade, k=c(10,4)) +
                          ti(measurement.no, duration) +
                          ti(measurement.no, preceding.front, k=c(10,3)) +
                       # level 3: 2d smooth by terms
                          ti(measurement.no, decade, k=c(10,4), by=stress) +
                       # random smooths
                          s(measurement.no, foll.broad.f3, bs="fs", m=1, k=4) +
                          s(measurement.no, speaker, bs="fs", m=1, k=4) +
                          s(measurement.no, word, bs="fs", m=1, k=4),
                      dat=R.m, method="ML",
                      AR.start=R.m$start.event, rho=R.m.rho)
  )
  # saveRDS(R.m.mod.AR, "models/R.m.mod.AR")
  
  summary(R.m.mod.AR)
  
  system.time(
    R.m.mod.AR.min.decade <- bam(f3 ~ stress +
                       # level 1 smooth terms
                          s(measurement.no) +
                          #s(decade, k=4) +
                          s(duration) +
                          s(preceding.front, k=3) +
                       # level 2: by terms
                          s(measurement.no, by=stress) +
                          #s(decade, k=4, by=stress) +
                       # level 2: 2d smooths
                          #ti(measurement.no, decade, k=c(10,4)) +
                          ti(measurement.no, duration) +
                          ti(measurement.no, preceding.front, k=c(10,3)) +
                       # level 3: 2d smooth by terms
                          #ti(measurement.no, decade, k=c(10,4), by=stress) +
                       # random smooths
                          s(measurement.no, foll.broad.f3, bs="fs", m=1, k=4) +
                          s(measurement.no, speaker, bs="fs", m=1, k=4) +
                          s(measurement.no, word, bs="fs", m=1, k=4),
                      dat=R.m, method="ML",
                      AR.start=R.m$start.event, rho=R.m.rho)
  )
# saveRDS(R.m.mod.AR.min.decade, "models/R.m.mod.AR.min.decade")
}

R.m.mod.no.AR <- readRDS("models/R.m.mod.no.AR")
R.m.mod.AR <- readRDS("models/R.m.mod.AR")
R.m.mod.AR.min.decade <- readRDS("models/R.m.mod.AR.min.decade")

compareML(R.m.mod.AR, R.m.mod.AR.min.decade)
plot_smooth(R.m.mod.AR, view="measurement.no", plot_all="decade", 
            cond=list(stress="full"), rm.ranef=T)
plot_smooth(R.m.mod.AR, view="measurement.no", plot_all="decade", 
            cond=list(stress="schwa"), rm.ranef=T)
```

Further questions:

- overall effect of stress?
- significant change in shape?
- do unstressed / stressed syllables show different changes?
  - different shape change?
  - different overall change?

```{r}
if (refit_please) {

  # overall stress?

  system.time(
    R.m.mod.AR.min.stress <- bam(f3 ~ #stress +
                       # level 1 smooth terms
                          s(measurement.no) +
                          s(decade, k=4) +
                          s(duration) +
                          s(preceding.front, k=3) +
                       # level 2: by terms
                          #s(measurement.no, by=stress) +
                          #s(decade, k=4, by=stress) +
                       # level 2: 2d smooths
                          ti(measurement.no, decade, k=c(10,4)) +
                          ti(measurement.no, duration) +
                          ti(measurement.no, preceding.front, k=c(10,3)) +
                       # level 3: 2d smooth by terms
                          #ti(measurement.no, decade, k=c(10,4), by=stress) +
                       # random smooths
                          s(measurement.no, foll.broad.f3, bs="fs", m=1, k=4) +
                          s(measurement.no, speaker, bs="fs", m=1, k=4) +
                          s(measurement.no, word, bs="fs", m=1, k=4),
                      dat=R.m, method="ML",
                      AR.start=R.m$start.event, rho=R.m.rho)
  )
  # saveRDS(R.m.mod.AR.min.stress, "models/R.m.mod.AR.min.stress")

  # shape change?

  system.time(
    R.m.mod.AR.min.decade.shape <- bam(f3 ~ stress +
                       # level 1 smooth terms
                          s(measurement.no) +
                          s(decade, k=4) +
                          s(duration) +
                          s(preceding.front, k=3) +
                       # level 2: by terms
                          s(measurement.no, by=stress) +
                          s(decade, k=4, by=stress) +
                       # level 2: 2d smooths
                          #ti(measurement.no, decade, k=c(10,4)) +
                          ti(measurement.no, duration) +
                          ti(measurement.no, preceding.front, k=c(10,3)) +
                       # level 3: 2d smooth by terms
                          #ti(measurement.no, decade, k=c(10,4), by=stress) +
                       # random smooths
                          s(measurement.no, foll.broad.f3, bs="fs", m=1, k=4) +
                          s(measurement.no, speaker, bs="fs", m=1, k=4) +
                          s(measurement.no, word, bs="fs", m=1, k=4),
                      dat=R.m, method="ML",
                      AR.start=R.m$start.event, rho=R.m.rho)
  )
  # saveRDS(R.m.mod.AR.min.decade.shape, "models/R.m.mod.AR.min.decade.shape")

  # overall stress x decade?

  system.time(
    R.m.mod.AR.min.decade.stress <- bam(f3 ~ stress +
                       # level 1 smooth terms
                          s(measurement.no) +
                          s(decade, k=4) +
                          s(duration) +
                          s(preceding.front, k=3) +
                       # level 2: by terms
                          s(measurement.no, by=stress) +
                          #s(decade, k=4, by=stress) +
                       # level 2: 2d smooths
                          ti(measurement.no, decade, k=c(10,4)) +
                          ti(measurement.no, duration) +
                          ti(measurement.no, preceding.front, k=c(10,3)) +
                       # level 3: 2d smooth by terms
                          #ti(measurement.no, decade, k=c(10,4), by=stress) +
                       # random smooths
                          s(measurement.no, foll.broad.f3, bs="fs", m=1, k=4) +
                          s(measurement.no, speaker, bs="fs", m=1, k=4) +
                          s(measurement.no, word, bs="fs", m=1, k=4),
                      dat=R.m, method="ML",
                      AR.start=R.m$start.event, rho=R.m.rho)
  )
  # saveRDS(R.m.mod.AR.min.decade.stress, "models/R.m.mod.AR.min.decade.stress")
  
  # shape stress x decade?
  
  system.time(
    R.m.mod.AR.min.decade.stress.shape <- bam(f3 ~ stress +
                       # level 1 smooth terms
                          s(measurement.no) +
                          s(decade, k=4) +
                          s(duration) +
                          s(preceding.front, k=3) +
                       # level 2: by terms
                          s(measurement.no, by=stress) +
                          s(decade, k=4, by=stress) +
                       # level 2: 2d smooths
                          ti(measurement.no, decade, k=c(10,4)) +
                          ti(measurement.no, duration) +
                          ti(measurement.no, preceding.front, k=c(10,3)) +
                       # level 3: 2d smooth by terms
                          #ti(measurement.no, decade, k=c(10,4), by=stress) +
                       # random smooths
                          s(measurement.no, foll.broad.f3, bs="fs", m=1, k=4) +
                          s(measurement.no, speaker, bs="fs", m=1, k=4) +
                          s(measurement.no, word, bs="fs", m=1, k=4),
                      dat=R.m, method="ML",
                      AR.start=R.m$start.event, rho=R.m.rho)
  )
  # saveRDS(R.m.mod.AR.min.decade.stress.shape, "models/R.m.mod.AR.min.decade.stress.shape")
}
  

R.m.mod.AR.min.stress <- readRDS("models/R.m.mod.AR.min.stress")
R.m.mod.AR.min.decade.shape <- readRDS("models/R.m.mod.AR.min.decade.shape")
R.m.mod.AR.min.decade.stress <- readRDS("models/R.m.mod.AR.min.decade.stress")
R.m.mod.AR.min.decade.stress.shape <- readRDS("models/R.m.mod.AR.min.decade.stress.shape")
# overall stress?
compareML(R.m.mod.AR, R.m.mod.AR.min.stress, suggest.report=T) # non-sig
# shape change?
compareML(R.m.mod.AR, R.m.mod.AR.min.decade.shape, suggest.report=T) # non-sig
# overall stress x decade?
compareML(R.m.mod.AR, R.m.mod.AR.min.decade.stress, suggest.report=T) # non-sig
# shape stress x decade?
compareML(R.m.mod.AR, R.m.mod.AR.min.decade.stress.shape) # non-sig
```

Make a pretty graph for the male predictions.

```{r}
R.m.full.plot <- plot_smooth(R.m.mod.AR, view="measurement.no", plot_all="decade", 
            cond=list(stress="full"), rm.ranef=T, n.grid=100)$fv
R.m.schwa.plot <- plot_smooth(R.m.mod.AR, view="measurement.no", plot_all="decade", 
            cond=list(stress="schwa"), rm.ranef=T, n.grid=100)$fv

R.m.plot <- rbind(R.m.full.plot, R.m.schwa.plot)
R.m.plot$decade <- factor(R.m.plot$decade, levels=c("1970","1980","1990","2000"))
  
ggplot(R.m.plot, aes(x=measurement.no, y=fit, group=decade, col=decade, lty=decade)) +
  geom_line(lwd=1) +
  facet_grid(.~stress) +
  geom_ribbon(aes(ymin=ll, ymax=ul, group=decade), col=NA,col="grey", alpha=0.1) +
  #scale_color_manual(name="sex",breaks=c("F","M"), labels=c("female", "male"), 
  #                   values=c("deepskyblue4","firebrick3")) +
  #scale_fill_manual(name="sex",breaks=c("F","M"), labels=c("female", "male"), 
  #                   values=c("deepskyblue4","firebrick3"), guide=F) +
  #scale_x_discrete(name="decade of recording", labels=c(1970,1980,1990,2000), expand=expand_scale(add=c(0.05,1))) +
  scale_color_manual(values=c("orange","darkorange3","deepskyblue1","deepskyblue4"),
                     name="decade of\nrecording") +
  scale_linetype_discrete(name="decade of\nrecording") +
  scale_x_continuous(name="measurement point", limits = c(0, 10), breaks=seq(0,10,2)) +
  scale_y_continuous(name="F3 (Hz)", limits=c(2000, 3050)) +
  theme_bw() +
  theme(axis.title = element_text(size=14), 
        axis.text = element_text(size=12),
        legend.title = element_text(size=14, face="bold"),
        legend.text=element_text(size=12),
        plot.title=element_text(size=14, face="bold"),
        panel.grid=element_blank(),
        strip.text=element_text(size=12)) +
  ggtitle("F3 changes for /r/ in males (not controlling for baseline F3)")
ggsave("graphs/r_male_f3_change_no_control.pdf", width=8, height=4)
```



# Females without baseline control

Fitting:

- model without AR for estimating rho
- full model with AR
- nested model with AR (rho from full model)

```{r}
if (refit_please) {
  system.time(
    R.f.mod.no.AR <- bam(f3 ~ stress +
                       # level 1 smooth terms
                          s(measurement.no) +
                          s(decade, k=4) +
                          s(duration) +
                          s(preceding.front, k=3) +
                       # level 2: by terms
                          s(measurement.no, by=stress) +
                          s(decade, k=4, by=stress) +
                       # level 2: 2d smooths
                          ti(measurement.no, decade, k=c(10,4)) +
                          ti(measurement.no, duration) +
                          ti(measurement.no, preceding.front, k=c(10,3)) +
                       # level 3: 2d smooth by terms
                          ti(measurement.no, decade, k=c(10,4), by=stress) +
                       # random smooths
                          s(measurement.no, foll.broad.f3, bs="fs", m=1, k=4) +
                          s(measurement.no, speaker, bs="fs", m=1, k=4) +
                          s(measurement.no, word, bs="fs", m=1, k=4),
                      dat=R.f)
  )
  # saveRDS(R.f.mod.no.AR, "models/R.f.mod.no.AR")
  
  R.f.rho <- start_value_rho(R.f.mod.no.AR)
  
  system.time(
    R.f.mod.AR <- bam(f3 ~ stress +
                       # level 1 smooth terms
                          s(measurement.no) +
                          s(decade, k=4) +
                          s(duration) +
                          s(preceding.front, k=3) +
                       # level 2: by terms
                          s(measurement.no, by=stress) +
                          s(decade, k=4, by=stress) +
                       # level 2: 2d smooths
                          ti(measurement.no, decade, k=c(10,4)) +
                          ti(measurement.no, duration) +
                          ti(measurement.no, preceding.front, k=c(10,3)) +
                       # level 3: 2d smooth by terms
                          ti(measurement.no, decade, k=c(10,4), by=stress) +
                       # random smooths
                          s(measurement.no, foll.broad.f3, bs="fs", m=1, k=4) +
                          s(measurement.no, speaker, bs="fs", m=1, k=4) +
                          s(measurement.no, word, bs="fs", m=1, k=4),
                      dat=R.f, method="ML",
                      AR.start=R.f$start.event, rho=R.f.rho)
  )
  # saveRDS(R.f.mod.AR, "models/R.f.mod.AR")
  
  summary(R.f.mod.AR)
  
  system.time(
    R.f.mod.AR.min.decade <- bam(f3 ~ stress +
                       # level 1 smooth terms
                          s(measurement.no) +
                          #s(decade, k=4) +
                          s(duration) +
                          s(preceding.front, k=3) +
                       # level 2: by terms
                          s(measurement.no, by=stress) +
                          #s(decade, k=4, by=stress) +
                       # level 2: 2d smooths
                          #ti(measurement.no, decade, k=c(10,4)) +
                          ti(measurement.no, duration) +
                          ti(measurement.no, preceding.front, k=c(10,3)) +
                       # level 3: 2d smooth by terms
                          #ti(measurement.no, decade, k=c(10,4), by=stress) +
                       # random smooths
                          s(measurement.no, foll.broad.f3, bs="fs", m=1, k=4) +
                          s(measurement.no, speaker, bs="fs", m=1, k=4) +
                          s(measurement.no, word, bs="fs", m=1, k=4),
                      dat=R.f, method="ML",
                      AR.start=R.f$start.event, rho=R.f.rho)
  )
  # saveRDS(R.f.mod.AR.min.decade, "models/R.f.mod.AR.min.decade")
}

R.f.mod.no.AR <- readRDS("models/R.f.mod.no.AR")
R.f.mod.AR <- readRDS("models/R.f.mod.AR")
R.f.mod.AR.min.decade <- readRDS("models/R.f.mod.AR.min.decade")

compareML(R.f.mod.AR, R.f.mod.AR.min.decade)
plot_smooth(R.f.mod.AR, view="measurement.no", plot_all="decade", 
            cond=list(stress="full"), rm.ranef=T)
plot_smooth(R.f.mod.AR, view="measurement.no", plot_all="decade", 
            cond=list(stress="schwa"), rm.ranef=T)
```

Further questions:

- overall effect of stress?
- significant change in shape?
- do unstressed / stressed syllables show different changes?
  - different shape change?
  - different overall change?

```{r}
if (refit_please) {

  # overall stress?

  system.time(
    R.f.mod.AR.min.stress <- bam(f3 ~ #stress +
                       # level 1 smooth terms
                          s(measurement.no) +
                          s(decade, k=4) +
                          s(duration) +
                          s(preceding.front, k=3) +
                       # level 2: by terms
                          #s(measurement.no, by=stress) +
                          #s(decade, k=4, by=stress) +
                       # level 2: 2d smooths
                          ti(measurement.no, decade, k=c(10,4)) +
                          ti(measurement.no, duration) +
                          ti(measurement.no, preceding.front, k=c(10,3)) +
                       # level 3: 2d smooth by terms
                          #ti(measurement.no, decade, k=c(10,4), by=stress) +
                       # random smooths
                          s(measurement.no, foll.broad.f3, bs="fs", m=1, k=4) +
                          s(measurement.no, speaker, bs="fs", m=1, k=4) +
                          s(measurement.no, word, bs="fs", m=1, k=4),
                      dat=R.f, method="ML",
                      AR.start=R.f$start.event, rho=R.f.rho)
  )
  # saveRDS(R.f.mod.AR.min.stress, "models/R.f.mod.AR.min.stress")

  # shape change?

  system.time(
    R.f.mod.AR.min.decade.shape <- bam(f3 ~ stress +
                       # level 1 smooth terms
                          s(measurement.no) +
                          s(decade, k=4) +
                          s(duration) +
                          s(preceding.front, k=3) +
                       # level 2: by terms
                          s(measurement.no, by=stress) +
                          s(decade, k=4, by=stress) +
                       # level 2: 2d smooths
                          #ti(measurement.no, decade, k=c(10,4)) +
                          ti(measurement.no, duration) +
                          ti(measurement.no, preceding.front, k=c(10,3)) +
                       # level 3: 2d smooth by terms
                          #ti(measurement.no, decade, k=c(10,4), by=stress) +
                       # random smooths
                          s(measurement.no, foll.broad.f3, bs="fs", m=1, k=4) +
                          s(measurement.no, speaker, bs="fs", m=1, k=4) +
                          s(measurement.no, word, bs="fs", m=1, k=4),
                      dat=R.f, method="ML",
                      AR.start=R.f$start.event, rho=R.f.rho)
  )
  # saveRDS(R.f.mod.AR.min.decade.shape, "models/R.f.mod.AR.min.decade.shape")

  # overall stress x decade?

  system.time(
    R.f.mod.AR.min.decade.stress <- bam(f3 ~ stress +
                       # level 1 smooth terms
                          s(measurement.no) +
                          s(decade, k=4) +
                          s(duration) +
                          s(preceding.front, k=3) +
                       # level 2: by terms
                          s(measurement.no, by=stress) +
                          #s(decade, k=4, by=stress) +
                       # level 2: 2d smooths
                          ti(measurement.no, decade, k=c(10,4)) +
                          ti(measurement.no, duration) +
                          ti(measurement.no, preceding.front, k=c(10,3)) +
                       # level 3: 2d smooth by terms
                          #ti(measurement.no, decade, k=c(10,4), by=stress) +
                       # random smooths
                          s(measurement.no, foll.broad.f3, bs="fs", m=1, k=4) +
                          s(measurement.no, speaker, bs="fs", m=1, k=4) +
                          s(measurement.no, word, bs="fs", m=1, k=4),
                      dat=R.f, method="ML",
                      AR.start=R.f$start.event, rho=R.f.rho)
  )
  # saveRDS(R.f.mod.AR.min.decade.stress, "models/R.f.mod.AR.min.decade.stress")
  
  # shape stress x decade?
  
  system.time(
    R.f.mod.AR.min.decade.stress.shape <- bam(f3 ~ stress +
                       # level 1 smooth terms
                          s(measurement.no) +
                          s(decade, k=4) +
                          s(duration) +
                          s(preceding.front, k=3) +
                       # level 2: by terms
                          s(measurement.no, by=stress) +
                          s(decade, k=4, by=stress) +
                       # level 2: 2d smooths
                          ti(measurement.no, decade, k=c(10,4)) +
                          ti(measurement.no, duration) +
                          ti(measurement.no, preceding.front, k=c(10,3)) +
                       # level 3: 2d smooth by terms
                          #ti(measurement.no, decade, k=c(10,4), by=stress) +
                       # random smooths
                          s(measurement.no, foll.broad.f3, bs="fs", m=1, k=4) +
                          s(measurement.no, speaker, bs="fs", m=1, k=4) +
                          s(measurement.no, word, bs="fs", m=1, k=4),
                      dat=R.f, method="ML",
                      AR.start=R.f$start.event, rho=R.f.rho)
  )
  # saveRDS(R.f.mod.AR.min.decade.stress.shape, "models/R.f.mod.AR.min.decade.stress.shape")
}
  

R.f.mod.AR.min.stress <- readRDS("models/R.f.mod.AR.min.stress")
R.f.mod.AR.min.decade.shape <- readRDS("models/R.f.mod.AR.min.decade.shape")
R.f.mod.AR.min.decade.stress <- readRDS("models/R.f.mod.AR.min.decade.stress")
R.f.mod.AR.min.decade.stress.shape <- readRDS("models/R.f.mod.AR.min.decade.stress.shape")
# overall stress?
compareML(R.f.mod.AR, R.f.mod.AR.min.stress, suggest.report=T) # sig
# shape change?
compareML(R.f.mod.AR, R.f.mod.AR.min.decade.shape, suggest.report=T) # sig
# overall stress x decade?
compareML(R.f.mod.AR, R.f.mod.AR.min.decade.stress, suggest.report=T) # non-sig
# shape stress x decade?
compareML(R.f.mod.AR, R.f.mod.AR.min.decade.stress.shape) # non-sig
```

Make a pretty graph for the female predictions.

```{r}
R.f.full.plot <- plot_smooth(R.f.mod.AR, view="measurement.no", plot_all="decade", 
            cond=list(stress="full"), rm.ranef=T, n.grid=100)$fv
R.f.schwa.plot <- plot_smooth(R.f.mod.AR, view="measurement.no", plot_all="decade", 
            cond=list(stress="schwa"), rm.ranef=T, n.grid=100)$fv

R.f.plot <- rbind(R.f.full.plot, R.f.schwa.plot)
R.f.plot$decade <- factor(R.f.plot$decade, levels=c("1970","1980","1990","2000"))
  
ggplot(R.f.plot, aes(x=measurement.no, y=fit, group=decade, col=decade, lty=decade)) +
  geom_line(lwd=1) +
  facet_grid(.~stress) +
  geom_ribbon(aes(ymin=ll, ymax=ul, group=decade), col=NA,col="grey", alpha=0.1) +
  #scale_color_manual(name="sex",breaks=c("F","M"), labels=c("female", "male"), 
  #                   values=c("deepskyblue4","firebrick3")) +
  #scale_fill_manual(name="sex",breaks=c("F","M"), labels=c("female", "male"), 
  #                   values=c("deepskyblue4","firebrick3"), guide=F) +
  #scale_x_discrete(name="decade of recording", labels=c(1970,1980,1990,2000), expand=expand_scale(add=c(0.05,1))) +
  scale_color_manual(name="decade of\nrecording", values=c("orange","darkorange3","deepskyblue1","deepskyblue4")) +
  scale_linetype_discrete(name="decade of\nrecording") +
  scale_x_continuous(name="measurement point", limits = c(0, 10), breaks=seq(0,10,2)) +
  scale_y_continuous(name="F3 (Hz)", limits=c(2000, 3050)) +
  theme_bw() +
  theme(axis.title = element_text(size=14), 
        axis.text = element_text(size=12),
        legend.title = element_text(size=14, face="bold"),
        legend.text=element_text(size=12),
        plot.title=element_text(size=14, face="bold"),
        panel.grid=element_blank(),
        strip.text=element_text(size=12)) +
  ggtitle("F3 changes for /r/ in females (not controlling for baseline F3)")
ggsave("graphs/r_female_f3_change_no_control.pdf", width=8, height=4)
```

# Males, take 2 (with baseline control)

```{r}
if (refit_please) {
  system.time(
    R.m.baseline.mod.no.AR <- bam(f3 ~ stress +
                          avg.f3 +
                       # level 1 smooth terms
                          s(measurement.no) +
                          s(decade, k=4) +
                          s(duration) +
                          s(preceding.front, k=3) +
                       # level 2: by terms
                          s(measurement.no, by=stress) +
                          s(decade, k=4, by=stress) +
                       # level 2: 2d smooths
                          ti(measurement.no, decade, k=c(10,4)) +
                          ti(measurement.no, duration) +
                          ti(measurement.no, preceding.front, k=c(10,3)) +
                       # level 3: 2d smooth by terms
                          ti(measurement.no, decade, k=c(10,4), by=stress) +
                       # random smooths
                          s(measurement.no, foll.broad.f3, bs="fs", m=1, k=4) +
                          s(measurement.no, speaker, bs="fs", m=1, k=4) +
                          s(measurement.no, word, bs="fs", m=1, k=4),
                      dat=R.m)
  )
  # saveRDS(R.m.baseline.mod.no.AR, "models/R.m.baseline.mod.no.AR")
  
  R.m.baseline.rho <- start_value_rho(R.m.baseline.mod.no.AR)
  
  system.time(
    R.m.baseline.mod.AR <- bam(f3 ~ stress +
                          avg.f3 +
                       # level 1 smooth terms
                          s(measurement.no) +
                          s(decade, k=4) +
                          s(duration) +
                          s(preceding.front, k=3) +
                       # level 2: by terms
                          s(measurement.no, by=stress) +
                          s(decade, k=4, by=stress) +
                       # level 2: 2d smooths
                          ti(measurement.no, decade, k=c(10,4)) +
                          ti(measurement.no, duration) +
                          ti(measurement.no, preceding.front, k=c(10,3)) +
                       # level 3: 2d smooth by terms
                          ti(measurement.no, decade, k=c(10,4), by=stress) +
                       # random smooths
                          s(measurement.no, foll.broad.f3, bs="fs", m=1, k=4) +
                          s(measurement.no, speaker, bs="fs", m=1, k=4) +
                          s(measurement.no, word, bs="fs", m=1, k=4),
                      dat=R.m, method="ML",
                      AR.start=R.m$start.event, rho=R.m.baseline.rho)
  )
  # saveRDS(R.m.baseline.mod.AR, "models/R.m.baseline.mod.AR")
  
  summary(R.m.baseline.mod.AR)
  
  system.time(
    R.m.baseline.mod.AR.min.decade <- bam(f3 ~ stress +
                          avg.f3 +
                       # level 1 smooth terms
                          s(measurement.no) +
                          #s(decade, k=4) +
                          s(duration) +
                          s(preceding.front, k=3) +
                       # level 2: by terms
                          s(measurement.no, by=stress) +
                          #s(decade, k=4, by=stress) +
                       # level 2: 2d smooths
                          #ti(measurement.no, decade, k=c(10,4)) +
                          ti(measurement.no, duration) +
                          ti(measurement.no, preceding.front, k=c(10,3)) +
                       # level 3: 2d smooth by terms
                          #ti(measurement.no, decade, k=c(10,4), by=stress) +
                       # random smooths
                          s(measurement.no, foll.broad.f3, bs="fs", m=1, k=4) +
                          s(measurement.no, speaker, bs="fs", m=1, k=4) +
                          s(measurement.no, word, bs="fs", m=1, k=4),
                      dat=R.m, method="ML",
                      AR.start=R.m$start.event, rho=R.m.baseline.rho)
  )
# saveRDS(R.m.baseline.mod.AR.min.decade, "models/R.m.baseline.mod.AR.min.decade")
}

R.m.baseline.mod.no.AR <- readRDS("models/R.m.baseline.mod.no.AR")
R.m.baseline.mod.AR <- readRDS("models/R.m.baseline.mod.AR")
R.m.baseline.mod.AR.min.decade <- readRDS("models/R.m.baseline.mod.AR.min.decade")

compareML(R.m.baseline.mod.AR, R.m.baseline.mod.AR.min.decade)
plot_smooth(R.m.baseline.mod.AR, view="measurement.no", plot_all="decade", 
            cond=list(stress="full"), rm.ranef=T)
plot_smooth(R.m.baseline.mod.AR, view="measurement.no", plot_all="decade", 
            cond=list(stress="schwa"), rm.ranef=T)
```

Further questions:

- overall effect of stress?
- significant change in shape?
- do unstressed / stressed syllables show different changes?
  - different shape change?
  - different overall change?

```{r}
if (refit_please) {

  # overall stress?

  system.time(
    R.m.baseline.mod.AR.min.stress <- bam(f3 ~ #stress +
                          avg.f3 +
                       # level 1 smooth terms
                          s(measurement.no) +
                          s(decade, k=4) +
                          s(duration) +
                          s(preceding.front, k=3) +
                       # level 2: by terms
                          #s(measurement.no, by=stress) +
                          #s(decade, k=4, by=stress) +
                       # level 2: 2d smooths
                          ti(measurement.no, decade, k=c(10,4)) +
                          ti(measurement.no, duration) +
                          ti(measurement.no, preceding.front, k=c(10,3)) +
                       # level 3: 2d smooth by terms
                          #ti(measurement.no, decade, k=c(10,4), by=stress) +
                       # random smooths
                          s(measurement.no, foll.broad.f3, bs="fs", m=1, k=4) +
                          s(measurement.no, speaker, bs="fs", m=1, k=4) +
                          s(measurement.no, word, bs="fs", m=1, k=4),
                      dat=R.m, method="ML",
                      AR.start=R.m$start.event, rho=R.m.baseline.rho)
  )
  # saveRDS(R.m.baseline.mod.AR.min.stress, "models/R.m.baseline.mod.AR.min.stress")

  # shape change?

  system.time(
    R.m.baseline.mod.AR.min.decade.shape <- bam(f3 ~ stress +
                          avg.f3 +
                       # level 1 smooth terms
                          s(measurement.no) +
                          s(decade, k=4) +
                          s(duration) +
                          s(preceding.front, k=3) +
                       # level 2: by terms
                          s(measurement.no, by=stress) +
                          s(decade, k=4, by=stress) +
                       # level 2: 2d smooths
                          #ti(measurement.no, decade, k=c(10,4)) +
                          ti(measurement.no, duration) +
                          ti(measurement.no, preceding.front, k=c(10,3)) +
                       # level 3: 2d smooth by terms
                          #ti(measurement.no, decade, k=c(10,4), by=stress) +
                       # random smooths
                          s(measurement.no, foll.broad.f3, bs="fs", m=1, k=4) +
                          s(measurement.no, speaker, bs="fs", m=1, k=4) +
                          s(measurement.no, word, bs="fs", m=1, k=4),
                      dat=R.m, method="ML",
                      AR.start=R.m$start.event, rho=R.m.baseline.rho)
  )
  # saveRDS(R.m.baseline.mod.AR.min.decade.shape, "models/R.m.baseline.mod.AR.min.decade.shape")

  # overall stress x decade?

  system.time(
    R.m.baseline.mod.AR.min.decade.stress <- bam(f3 ~ stress +
                          avg.f3 +
                       # level 1 smooth terms
                          s(measurement.no) +
                          s(decade, k=4) +
                          s(duration) +
                          s(preceding.front, k=3) +
                       # level 2: by terms
                          s(measurement.no, by=stress) +
                          #s(decade, k=4, by=stress) +
                       # level 2: 2d smooths
                          ti(measurement.no, decade, k=c(10,4)) +
                          ti(measurement.no, duration) +
                          ti(measurement.no, preceding.front, k=c(10,3)) +
                       # level 3: 2d smooth by terms
                          #ti(measurement.no, decade, k=c(10,4), by=stress) +
                       # random smooths
                          s(measurement.no, foll.broad.f3, bs="fs", m=1, k=4) +
                          s(measurement.no, speaker, bs="fs", m=1, k=4) +
                          s(measurement.no, word, bs="fs", m=1, k=4),
                      dat=R.m, method="ML",
                      AR.start=R.m$start.event, rho=R.m.baseline.rho)
  )
  # saveRDS(R.m.baseline.mod.AR.min.decade.stress, "models/R.m.baseline.mod.AR.min.decade.stress")
  
  # shape stress x decade?
  
  system.time(
    R.m.baseline.mod.AR.min.decade.stress.shape <- bam(f3 ~ stress +
                          avg.f3 +
                       # level 1 smooth terms
                          s(measurement.no) +
                          s(decade, k=4) +
                          s(duration) +
                          s(preceding.front, k=3) +
                       # level 2: by terms
                          s(measurement.no, by=stress) +
                          s(decade, k=4, by=stress) +
                       # level 2: 2d smooths
                          ti(measurement.no, decade, k=c(10,4)) +
                          ti(measurement.no, duration) +
                          ti(measurement.no, preceding.front, k=c(10,3)) +
                       # level 3: 2d smooth by terms
                          #ti(measurement.no, decade, k=c(10,4), by=stress) +
                       # random smooths
                          s(measurement.no, foll.broad.f3, bs="fs", m=1, k=4) +
                          s(measurement.no, speaker, bs="fs", m=1, k=4) +
                          s(measurement.no, word, bs="fs", m=1, k=4),
                      dat=R.m, method="ML",
                      AR.start=R.m$start.event, rho=R.m.baseline.rho)
  )
  # saveRDS(R.m.baseline.mod.AR.min.decade.stress.shape, "models/R.m.baseline.mod.AR.min.decade.stress.shape")
}
  

R.m.baseline.mod.AR.min.stress <- readRDS("models/R.m.baseline.mod.AR.min.stress")
R.m.baseline.mod.AR.min.decade.shape <- readRDS("models/R.m.baseline.mod.AR.min.decade.shape")
R.m.baseline.mod.AR.min.decade.stress <- readRDS("models/R.m.baseline.mod.AR.min.decade.stress")
R.m.baseline.mod.AR.min.decade.stress.shape <- readRDS("models/R.m.baseline.mod.AR.min.decade.stress.shape")
# overall stress?
compareML(R.m.baseline.mod.AR, R.m.baseline.mod.AR.min.stress, suggest.report=T) # non-sig
# shape change?
compareML(R.m.baseline.mod.AR, R.m.baseline.mod.AR.min.decade.shape, suggest.report=T) # non-sig
# overall stress x decade?
compareML(R.m.baseline.mod.AR, R.m.baseline.mod.AR.min.decade.stress, suggest.report=T) # non-sig
# shape stress x decade?
compareML(R.m.baseline.mod.AR, R.m.baseline.mod.AR.min.decade.stress.shape) # non-sig
```

Make a pretty graph for the male predictions.

```{r}
R.m.baseline.full.plot <- plot_smooth(R.m.baseline.mod.AR, view="measurement.no", plot_all="decade", 
            cond=list(stress="full"), rm.ranef=T, n.grid=100)$fv
R.m.baseline.schwa.plot <- plot_smooth(R.m.baseline.mod.AR, view="measurement.no", plot_all="decade", 
            cond=list(stress="schwa"), rm.ranef=T, n.grid=100)$fv

R.m.baseline.plot <- rbind(R.m.baseline.full.plot, R.m.baseline.schwa.plot)
R.m.baseline.plot$decade <- factor(R.m.baseline.plot$decade, levels=c("1970","1980","1990","2000"))
  
ggplot(R.m.baseline.plot, aes(x=measurement.no, y=fit, group=decade, col=decade, lty=decade)) +
  geom_line(lwd=1) +
  facet_grid(.~stress) +
  geom_ribbon(aes(ymin=ll, ymax=ul, group=decade), col=NA,col="grey", alpha=0.1) +
  #scale_color_manual(name="sex",breaks=c("F","M"), labels=c("female", "male"), 
  #                   values=c("deepskyblue4","firebrick3")) +
  #scale_fill_manual(name="sex",breaks=c("F","M"), labels=c("female", "male"), 
  #                   values=c("deepskyblue4","firebrick3"), guide=F) +
  #scale_x_discrete(name="decade of recording", labels=c(1970,1980,1990,2000), expand=expand_scale(add=c(0.05,1))) +
  scale_color_manual(name="decade of\nrecording", values=c("orange","darkorange3","deepskyblue1","deepskyblue4")) +
  scale_linetype_discrete(name="decade of\nrecording") +
  scale_x_continuous(name="measurement point", limits = c(0, 10), breaks=seq(0,10,2)) +
  scale_y_continuous(name="F3 (Hz)", limits=c(2000, 3050)) +
  theme_bw() +
  theme(axis.title = element_text(size=14), 
        axis.text = element_text(size=12),
        legend.title = element_text(size=14, face="bold"),
        legend.text=element_text(size=12),
        plot.title=element_text(size=14, face="bold"),
        panel.grid=element_blank(),
        strip.text=element_text(size=12)) +
  ggtitle("F3 changes for /r/ in males (controlling for baseline F3)")
ggsave("graphs/r_male_f3_change_control.pdf", width=8, height=4)
```



# Females, take 2 (with baseline control)

Fitting:

- model without AR for estimating rho
- full model with AR
- nested model with AR (rho from full model)

```{r}
if (refit_please) {
  system.time(
    R.f.baseline.mod.no.AR <- bam(f3 ~ stress +
                          avg.f3 +
                       # level 1 smooth terms
                          s(measurement.no) +
                          s(decade, k=4) +
                          s(duration) +
                          s(preceding.front, k=3) +
                       # level 2: by terms
                          s(measurement.no, by=stress) +
                          s(decade, k=4, by=stress) +
                       # level 2: 2d smooths
                          ti(measurement.no, decade, k=c(10,4)) +
                          ti(measurement.no, duration) +
                          ti(measurement.no, preceding.front, k=c(10,3)) +
                       # level 3: 2d smooth by terms
                          ti(measurement.no, decade, k=c(10,4), by=stress) +
                       # random smooths
                          s(measurement.no, foll.broad.f3, bs="fs", m=1, k=4) +
                          s(measurement.no, speaker, bs="fs", m=1, k=4) +
                          s(measurement.no, word, bs="fs", m=1, k=4),
                      dat=R.f)
  )
  # saveRDS(R.f.baseline.mod.no.AR, "models/R.f.baseline.mod.no.AR")
  
  R.f.baseline.rho <- start_value_rho(R.f.baseline.mod.no.AR)
  
  system.time(
    R.f.baseline.mod.AR <- bam(f3 ~ stress +
                          avg.f3 +
                       # level 1 smooth terms
                          s(measurement.no) +
                          s(decade, k=4) +
                          s(duration) +
                          s(preceding.front, k=3) +
                       # level 2: by terms
                          s(measurement.no, by=stress) +
                          s(decade, k=4, by=stress) +
                       # level 2: 2d smooths
                          ti(measurement.no, decade, k=c(10,4)) +
                          ti(measurement.no, duration) +
                          ti(measurement.no, preceding.front, k=c(10,3)) +
                       # level 3: 2d smooth by terms
                          ti(measurement.no, decade, k=c(10,4), by=stress) +
                       # random smooths
                          s(measurement.no, foll.broad.f3, bs="fs", m=1, k=4) +
                          s(measurement.no, speaker, bs="fs", m=1, k=4) +
                          s(measurement.no, word, bs="fs", m=1, k=4),
                      dat=R.f, method="ML",
                      AR.start=R.f$start.event, rho=R.f.baseline.rho)
  )
  # saveRDS(R.f.baseline.mod.AR, "models/R.f.baseline.mod.AR")
  
  summary(R.f.baseline.mod.AR)
  
  system.time(
    R.f.baseline.mod.AR.min.decade <- bam(f3 ~ stress +
                          avg.f3 +
                       # level 1 smooth terms
                          s(measurement.no) +
                          #s(decade, k=4) +
                          s(duration) +
                          s(preceding.front, k=3) +
                       # level 2: by terms
                          s(measurement.no, by=stress) +
                          #s(decade, k=4, by=stress) +
                       # level 2: 2d smooths
                          #ti(measurement.no, decade, k=c(10,4)) +
                          ti(measurement.no, duration) +
                          ti(measurement.no, preceding.front, k=c(10,3)) +
                       # level 3: 2d smooth by terms
                          #ti(measurement.no, decade, k=c(10,4), by=stress) +
                       # random smooths
                          s(measurement.no, foll.broad.f3, bs="fs", m=1, k=4) +
                          s(measurement.no, speaker, bs="fs", m=1, k=4) +
                          s(measurement.no, word, bs="fs", m=1, k=4),
                      dat=R.f, method="ML",
                      AR.start=R.f$start.event, rho=R.f.baseline.rho)
  )
  # saveRDS(R.f.baseline.mod.AR.min.decade, "models/R.f.baseline.mod.AR.min.decade")
}

R.f.baseline.mod.no.AR <- readRDS("models/R.f.baseline.mod.no.AR")
R.f.baseline.mod.AR <- readRDS("models/R.f.baseline.mod.AR")
R.f.baseline.mod.AR.min.decade <- readRDS("models/R.f.baseline.mod.AR.min.decade")

compareML(R.f.baseline.mod.AR, R.f.baseline.mod.AR.min.decade)
plot_smooth(R.f.baseline.mod.AR, view="measurement.no", plot_all="decade", 
            cond=list(stress="full"), rm.ranef=T)
plot_smooth(R.f.baseline.mod.AR, view="measurement.no", plot_all="decade", 
            cond=list(stress="schwa"), rm.ranef=T)
```

Further questions:

- overall effect of stress?
- significant change in shape?
- do unstressed / stressed syllables show different changes?
  - different shape change?
  - different overall change?

```{r}
if (refit_please) {

  # overall stress?

  system.time(
    R.f.baseline.mod.AR.min.stress <- bam(f3 ~ #stress +
                          avg.f3 +
                       # level 1 smooth terms
                          s(measurement.no) +
                          s(decade, k=4) +
                          s(duration) +
                          s(preceding.front, k=3) +
                       # level 2: by terms
                          #s(measurement.no, by=stress) +
                          #s(decade, k=4, by=stress) +
                       # level 2: 2d smooths
                          ti(measurement.no, decade, k=c(10,4)) +
                          ti(measurement.no, duration) +
                          ti(measurement.no, preceding.front, k=c(10,3)) +
                       # level 3: 2d smooth by terms
                          #ti(measurement.no, decade, k=c(10,4), by=stress) +
                       # random smooths
                          s(measurement.no, foll.broad.f3, bs="fs", m=1, k=4) +
                          s(measurement.no, speaker, bs="fs", m=1, k=4) +
                          s(measurement.no, word, bs="fs", m=1, k=4),
                      dat=R.f, method="ML",
                      AR.start=R.f$start.event, rho=R.f.baseline.rho)
  )
  saveRDS(R.f.baseline.mod.AR.min.stress, "models/R.f.baseline.mod.AR.min.stress")

  # shape change?

  system.time(
    R.f.baseline.mod.AR.min.decade.shape <- bam(f3 ~ stress +
                          avg.f3 +
                       # level 1 smooth terms
                          s(measurement.no) +
                          s(decade, k=4) +
                          s(duration) +
                          s(preceding.front, k=3) +
                       # level 2: by terms
                          s(measurement.no, by=stress) +
                          s(decade, k=4, by=stress) +
                       # level 2: 2d smooths
                          #ti(measurement.no, decade, k=c(10,4)) +
                          ti(measurement.no, duration) +
                          ti(measurement.no, preceding.front, k=c(10,3)) +
                       # level 3: 2d smooth by terms
                          #ti(measurement.no, decade, k=c(10,4), by=stress) +
                       # random smooths
                          s(measurement.no, foll.broad.f3, bs="fs", m=1, k=4) +
                          s(measurement.no, speaker, bs="fs", m=1, k=4) +
                          s(measurement.no, word, bs="fs", m=1, k=4),
                      dat=R.f, method="ML",
                      AR.start=R.f$start.event, rho=R.f.baseline.rho)
  )
  saveRDS(R.f.baseline.mod.AR.min.decade.shape, "models/R.f.baseline.mod.AR.min.decade.shape")

  # overall stress x decade?

  system.time(
    R.f.baseline.mod.AR.min.decade.stress <- bam(f3 ~ stress +
                          avg.f3 +
                       # level 1 smooth terms
                          s(measurement.no) +
                          s(decade, k=4) +
                          s(duration) +
                          s(preceding.front, k=3) +
                       # level 2: by terms
                          s(measurement.no, by=stress) +
                          #s(decade, k=4, by=stress) +
                       # level 2: 2d smooths
                          ti(measurement.no, decade, k=c(10,4)) +
                          ti(measurement.no, duration) +
                          ti(measurement.no, preceding.front, k=c(10,3)) +
                       # level 3: 2d smooth by terms
                          #ti(measurement.no, decade, k=c(10,4), by=stress) +
                       # random smooths
                          s(measurement.no, foll.broad.f3, bs="fs", m=1, k=4) +
                          s(measurement.no, speaker, bs="fs", m=1, k=4) +
                          s(measurement.no, word, bs="fs", m=1, k=4),
                      dat=R.f, method="ML",
                      AR.start=R.f$start.event, rho=R.f.baseline.rho)
  )
  saveRDS(R.f.baseline.mod.AR.min.decade.stress, "models/R.f.baseline.mod.AR.min.decade.stress")
  
  # shape stress x decade?
  
  system.time(
    R.f.baseline.mod.AR.min.decade.stress.shape <- bam(f3 ~ stress +
                          avg.f3 +
                       # level 1 smooth terms
                          s(measurement.no) +
                          s(decade, k=4) +
                          s(duration) +
                          s(preceding.front, k=3) +
                       # level 2: by terms
                          s(measurement.no, by=stress) +
                          s(decade, k=4, by=stress) +
                       # level 2: 2d smooths
                          ti(measurement.no, decade, k=c(10,4)) +
                          ti(measurement.no, duration) +
                          ti(measurement.no, preceding.front, k=c(10,3)) +
                       # level 3: 2d smooth by terms
                          #ti(measurement.no, decade, k=c(10,4), by=stress) +
                       # random smooths
                          s(measurement.no, foll.broad.f3, bs="fs", m=1, k=4) +
                          s(measurement.no, speaker, bs="fs", m=1, k=4) +
                          s(measurement.no, word, bs="fs", m=1, k=4),
                      dat=R.f, method="ML",
                      AR.start=R.f$start.event, rho=R.f.baseline.rho)
  )
  saveRDS(R.f.baseline.mod.AR.min.decade.stress.shape, "models/R.f.baseline.mod.AR.min.decade.stress.shape")
}
  

R.f.baseline.mod.AR.min.stress <- readRDS("models/R.f.baseline.mod.AR.min.stress")
R.f.baseline.mod.AR.min.decade.shape <- readRDS("models/R.f.baseline.mod.AR.min.decade.shape")
R.f.baseline.mod.AR.min.decade.stress <- readRDS("models/R.f.baseline.mod.AR.min.decade.stress")
R.f.baseline.mod.AR.min.decade.stress.shape <- readRDS("models/R.f.baseline.mod.AR.min.decade.stress.shape")
# overall stress?
compareML(R.f.baseline.mod.AR, R.f.baseline.mod.AR.min.stress, suggest.report=T) # non-sig
# shape change?
compareML(R.f.baseline.mod.AR, R.f.baseline.mod.AR.min.decade.shape, suggest.report=T) # non-sig
# overall stress x decade?
compareML(R.f.baseline.mod.AR, R.f.baseline.mod.AR.min.decade.stress, suggest.report=T) # non-sig
# shape stress x decade?
compareML(R.f.baseline.mod.AR, R.f.baseline.mod.AR.min.decade.stress.shape) # non-sig
```

Make a pretty graph for the female predictions.

```{r}
R.f.baseline.full.plot <- plot_smooth(R.f.baseline.mod.AR, view="measurement.no", plot_all="decade", 
            cond=list(stress="full"), rm.ranef=T, n.grid=100)$fv
R.f.baseline.schwa.plot <- plot_smooth(R.f.baseline.mod.AR, view="measurement.no", plot_all="decade", 
            cond=list(stress="schwa"), rm.ranef=T, n.grid=100)$fv

R.f.baseline.plot <- rbind(R.f.baseline.full.plot, R.f.baseline.schwa.plot)
R.f.baseline.plot$decade <- factor(R.f.baseline.plot$decade, levels=c("1970","1980","1990","2000"))
  
ggplot(R.f.baseline.plot, aes(x=measurement.no, y=fit, group=decade, col=decade, lty=decade)) +
  geom_line(lwd=1) +
  facet_grid(.~stress) +
  geom_ribbon(aes(ymin=ll, ymax=ul, group=decade), col=NA,col="grey", alpha=0.1) +
  #scale_color_manual(name="sex",breaks=c("F","M"), labels=c("female", "male"), 
  #                   values=c("deepskyblue4","firebrick3")) +
  #scale_fill_manual(name="sex",breaks=c("F","M"), labels=c("female", "male"), 
  #                   values=c("deepskyblue4","firebrick3"), guide=F) +
  #scale_x_discrete(name="decade of recording", labels=c(1970,1980,1990,2000), expand=expand_scale(add=c(0.05,1))) +
  scale_color_manual(name="decade of\nrecording", values=c("orange","darkorange3","deepskyblue1","deepskyblue4")) +
  scale_linetype_discrete(name="decade of\nrecording") +
  scale_x_continuous(name="measurement point", limits = c(0, 10), breaks=seq(0,10,2)) +
  scale_y_continuous(name="F3 (Hz)", limits=c(2000, 3050)) +
  theme_bw() +
  theme(axis.title = element_text(size=14), 
        axis.text = element_text(size=12),
        legend.title = element_text(size=14, face="bold"),
        legend.text=element_text(size=12),
        plot.title=element_text(size=14, face="bold"),
        panel.grid=element_blank(),
        strip.text=element_text(size=12)) +
  ggtitle("F3 changes for /r/ in females (controlling for baseline F3)")
ggsave("graphs/r_female_f3_change_control.pdf", width=8, height=4)
```

## Auditory analysis of R realisations

```{r}
# create data set with unique realisations
R.aud <- R[!(R$trans %in% c("", "?r")) & R$measurement.no==0,]
# only a few weakened approximants, so fold these into other categories
R.aud <- R %>%
  filter(!(trans %in% c("", "?r")), 
         !is.na(trans),
         measurement.no==0) %>%
  mutate(trans.broad=recode(trans, wa="a"),
         gender=recode(gender, M="male", `F`="female"))

# by decade / gender graph
R.props.dec <- R.aud %>% 
  count(decade, gender, trans.broad) %>% 
  group_by(decade, gender) %>%
  mutate(prop = n/sum(n)) %>%
  ungroup()

R.props.dec$trans.broad <- factor(R.props.dec$trans.broad, levels=c("o","i","wt","a","t","r"))
R.props.dec
ggplot(R.props.dec, aes(x=decade, fill=trans.broad)) + 
  geom_bar(aes(y=prop), stat="identity", position="stack") + 
  facet_wrap(~gender) + 
  scale_fill_manual(values = c("orange","darkorange1",
                               "darkorange3","deepskyblue1",
                               "deepskyblue3","deepskyblue4"),
                     name="/r/ realisation", 
                    labels=c("zero","intermediate",
                             "weakened tap","approximant",
                             "tap","trill")) +
  scale_y_continuous(labels = scales::percent, name="percentage of realisations") +
  scale_x_continuous(name = "decade of recording") +
  theme_bw() +
  theme(axis.title = element_text(size=14), 
        axis.text = element_text(size=12),
        legend.title = element_text(size=14, face="bold"),
        legend.text=element_text(size=12),
        plot.title=element_text(size=14, face="bold"),
        panel.grid=element_blank(),
        strip.text=element_text(size=12)) +
ggtitle("Changes in proportions of auditory /r/ variants")
ggsave("graphs/r_auditory.pdf", width=8, height=4)
```

Let's run some logistic regression models...

```{r}
R.aud <- mutate(R.aud, r.strength=recode(trans.broad, 
                                    t="strong",
                                    a="strong",
                                    r="strong",
                                    wt="weak",
                                    i="weak",
                                    o="weak"),
                       r.presence=recode(trans.broad, 
                                    t="present",
                                    a="present",
                                    r="present",
                                    wt="present",
                                    i="deleted",
                                    o="deleted") 
                )
R.aud$r.strength <- factor(R.aud$r.strength, levels=c("weak","strong"))
R.aud$r.presence <- factor(R.aud$r.presence, levels=c("deleted","present"))
if (refit_please) {
  r.str.mod <- glmer(r.strength ~ scale(decade) * gender + stress +
                                (1 | speaker) +
                                (1 | word),
                        data=R.aud,
                        family="binomial")
  #saveRDS(r.str.mod, "models/r.str.mod")
  
  r.str.mod.min.decade <- glmer(r.strength ~ gender + stress +
                                (1 | speaker) +
                                (1 | word),
                        data=R.aud,
                        family="binomial")
  #saveRDS(r.str.mod.min.decade, "models/r.str.mod.min.decade")
  
  r.pres.mod <- glmer(r.presence ~ scale(decade) * gender + stress +
                                (1 | speaker) +
                                (1 | word),
                        data=R.aud,
                        family="binomial",
                        control=glmerControl(optimizer="bobyqa",
                              optCtrl=list(maxfun=2e5)))
  #saveRDS(r.pres.mod, "models/r.pres.mod")
  
  r.pres.mod.min.decade <- glmer(r.presence ~ gender + stress +
                                (1 | speaker) +
                                (1 | word),
                        data=R.aud,
                        family="binomial")
  #saveRDS(r.pres.mod.min.decade, "models/r.pres.mod.min.decade")
}

r.str.mod <- readRDS("models/r.str.mod")
r.str.mod.min.decade <- readRDS("models/r.str.mod.min.decade")
r.pres.mod <- readRDS("models/r.pres.mod")
r.pres.mod.min.decade <- readRDS("models/r.pres.mod.min.decade")

summary(r.str.mod)
summary(r.pres.mod)

anova(r.str.mod, r.str.mod.min.decade,test="Chisq")
anova(r.pres.mod, r.pres.mod.min.decade,test="Chisq")
```

Nothing really significant here, and these are slightly anti-conservative models, so no point in pushing this any further.

How about the phonetics of different R variants? Modelling F3 as a function of variant type should do the trick...

```{r}
R <- mutate(R, trans.broad=recode(trans, wa="a"))
R <- R[!(R$trans %in% c("", "?r")),]
R$trans.broad.fact <- as.ordered(R$trans.broad)
contrasts(R$trans.broad.fact) <- "contr.treatment"
R$gender <- as.ordered(R$gender)
contrasts(R$gender) <- "contr.treatment"

if (refit_please) {

  system.time(
    R.shapes.mod.no.AR <- bam(f3 ~ stress + trans.broad.fact + gender +
                       # level 1 smooth terms
                          s(measurement.no) +
                          s(duration) +
                          s(preceding.front, k=3) +
                       # level 2: by terms
                          s(measurement.no, by=stress) +
                          s(measurement.no, by=trans.broad.fact) +
                          s(measurement.no, by=gender) +
                       # level 2: 2d smooths
                          ti(measurement.no, duration) +
                          ti(measurement.no, preceding.front, k=c(10,3)) +
                       # random smooths
                          s(measurement.no, foll.broad.f3, bs="fs", m=1, k=4) +
                          s(measurement.no, speaker, bs="fs", m=1, k=4) +
                          s(measurement.no, word, bs="fs", m=1, k=4),
                      dat=R)
  )
  saveRDS(R.shapes.mod.no.AR, "models/R.shapes.mod.no.AR")
  
  R.shapes.rho <- start_value_rho(R.shapes.mod.no.AR)
  R$start.event <- R$measurement.no == 0
  
  system.time(
    R.shapes.mod.AR   <- bam(f3 ~ stress + trans.broad.fact + gender +
                       # level 1 smooth terms
                          s(measurement.no) +
                          s(duration) +
                          s(preceding.front, k=3) +
                       # level 2: by terms
                          s(measurement.no, by=stress) +
                          s(measurement.no, by=trans.broad.fact) +
                          s(measurement.no, by=gender) +
                       # level 2: 2d smooths
                          ti(measurement.no, duration) +
                          ti(measurement.no, preceding.front, k=c(10,3)) +
                       # random smooths
                          s(measurement.no, foll.broad.f3, bs="fs", m=1, k=4) +
                          s(measurement.no, speaker, bs="fs", m=1, k=4) +
                          s(measurement.no, word, bs="fs", m=1, k=4),
                      dat=R,
                      AR.start=R$start.event, rho=R.shapes.rho,
                      method="ML")
  )
  saveRDS(R.shapes.mod.AR, "models/R.shapes.mod.AR")
  
  summary(R.shapes.mod.AR)
  
  system.time(
    R.shapes.mod.AR.min.trans   <- bam(f3 ~ stress + gender + # trans.broad.fact + 
                       # level 1 smooth terms
                          s(measurement.no) +
                          s(duration) +
                          s(preceding.front, k=3) +
                       # level 2: by terms
                          s(measurement.no, by=stress) +
                          #s(measurement.no, by=trans.broad.fact) +
                          s(measurement.no, by=gender) +
                       # level 2: 2d smooths
                          ti(measurement.no, duration) +
                          ti(measurement.no, preceding.front, k=c(10,3)) +
                       # random smooths
                          s(measurement.no, foll.broad.f3, bs="fs", m=1, k=4) +
                          s(measurement.no, speaker, bs="fs", m=1, k=4) +
                          s(measurement.no, word, bs="fs", m=1, k=4),
                      dat=R,
                      AR.start=R$start.event, rho=R.shapes.rho,
                      method="ML")
  )
  saveRDS(R.shapes.mod.AR.min.trans, "models/R.shapes.mod.AR.min.trans")
}

#summary(R.shapes.mod.AR.min.trans)


R.shapes.mod.no.AR <- readRDS("models/R.shapes.mod.no.AR")
R.shapes.mod.AR <- readRDS("models/R.shapes.mod.AR")
R.shapes.mod.AR.min.trans <- readRDS("models/R.shapes.mod.AR.min.trans")

compareML(R.shapes.mod.AR, R.shapes.mod.AR.min.trans)

#plot_smooth(R.shapes.mod.AR, view="measurement.no", plot_all="trans.broad.fact",
#            rm.ranef=T, rug=F, se=0)
```

Pretty plot for paper:

```{r}
R.shapes <- plot_smooth(R.shapes.mod.AR, view="measurement.no", plot_all="trans.broad.fact",
                        cond=list(gender="F"), rm.ranef=T, rug=F)[[1]]

R.shapes$trans.broad.fact <- factor(R.shapes$trans.broad.fact, levels=c("o","i","wt","a","t","r"))
R.shapes <- R.shapes %>% 
  mutate(realisation=recode(trans.broad.fact,
                            o="zero",
                            i="intermediate",
                            wt="weakened tap",
                            a="approximant",
                            t="full tap",
                            r="trill"),
         realisation=factor(realisation, 
                            levels=c("zero",
                                     "intermediate",
                                     "weakened tap",
                                     "approximant",
                                     "full tap",
                                     "trill"))
  )
R.shapes <- filter(R.shapes, trans.broad.fact != "r")

ggplot(R.shapes, aes(x=measurement.no, y=fit, col=realisation, lty=realisation)) +
  geom_ribbon(aes(ymin=ll, ymax=ul, group=trans.broad.fact), alpha=0.05, colour=NA) +
  geom_line(lwd=1) +
  scale_linetype_manual(values = c(3,4,2,
                              5,1),
                     name="/r/ realisation") +
  scale_colour_manual(values = c("orange","darkorange1",
                               "darkorange3","deepskyblue1",
                               "deepskyblue3"),
                      name="/r/ realisation") +
  scale_x_continuous(name="measurement point", limits = c(0, 10), breaks=seq(0,10,2)) +
  scale_y_continuous(name="F3 (Hz)", limits=c(2450,2950)) +
  theme_bw() +
  theme(axis.title = element_text(size=14), 
        axis.text = element_text(size=12),
        legend.title = element_text(size=14, face="bold"),
        legend.text=element_text(size=12),
        plot.title=element_text(size=14, face="bold"),
        panel.grid=element_blank(),
        strip.text=element_text(size=12)) +
  ggtitle("Acoustic correlates of different /r/ realisations in F3")
ggsave("graphs/r_acoustic_correlates.pdf", width=6, height=4)
```



## Auditory VQ analysis

```{r}
avg.f3 <- unique(select(R, speaker, avg.f3))
R.vq <- read_csv("data/auditoryvq.csv") %>%
  rename(speaker="speaker code",
         advanced.tip.blade="advanced tip/blade",
         tongue.body.height="tongue body height",
         tongue.body.front.backness="tongue body front-backness") %>%
  inner_join(avg.f3, by="speaker") %>%
  mutate(gender=as.factor(ifelse(!is.na(str_match(speaker, "-m0")), "male", "female")),
         decade=unlist(substr(speaker, 1, 2)),
         decade=as.numeric(recode(decade, `00`="100")) + 1900)


# F3

vq.f3.model <- lm(avg.f3 ~ advanced.tip.blade + 
             tongue.body.height + 
             tongue.body.front.backness + 
             gender, data=R.vq)
summary(vq.f3.model)



# overall VQ

R.vq$decade.centred <- scale(R.vq$decade, scale=F)
R.vq$gender.sum <- R.vq$gender
contrasts(R.vq$gender.sum) <- contr.sum(2)

summary(lm(tongue.body.height ~ decade.centred + gender.sum, data=R.vq))
summary(lm(tongue.body.front.backness ~ decade.centred + gender.sum, data=R.vq))
summary(lm(advanced.tip.blade ~ decade.centred + gender.sum, data=R.vq))


R.vq.count <- dplyr::count(R.vq, tongue.body.height, decade)

R.vq.ordered <- R.vq %>% 
  arrange(decade, tongue.body.height)
ggplot(R.vq.ordered, aes(x=factor(decade), y=tongue.body.height)) +
  geom_point(x=1,y=1, aes(colour="Female")) +
  geom_point(x=1,y=1, aes(colour="Male")) +
  scale_color_manual(name="Gender", values=c("deepskyblue4","orange"), breaks=c("Female","Male")) +
  geom_dotplot(binaxis="y", stackdir="center", binpositions="all",
               fill=ifelse(R.vq.ordered$gender=="female", "deepskyblue4","orange"),
               col=NA, dotsize=1.5) +
  stat_summary(fun.y=mean, fun.ymin=mean, fun.ymax=mean, geom="crossbar", aes(group=decade), width=0.4, col="grey") +
  scale_x_discrete(name="Decade of recording") +
  scale_y_continuous(name="Tongue body height") +
  guides(colour = guide_legend(override.aes = list(size = 4))) +
  theme_bw() +
  theme(axis.title = element_text(size=14), 
        axis.text = element_text(size=12),
        legend.title = element_text(size=14, face="bold"),
        legend.text=element_text(size=12),
        plot.title=element_text(size=14, face="bold"),
        panel.grid=element_blank(),
        strip.text=element_text(size=12)) +
  ggtitle("Changes in tongue body height over time")
ggsave("graphs/tongue_body_height.pdf", width=6, height=4)
```